---
title: "Bte_necklace_limma"
author: "Anna Orteu"
date: "24/05/2018"
output: 
  html_document:
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 4  # upto 4 depths of headings (specified by #, ##, ### and ####)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: flatly
    code_folding: hide # Add some awesome buttons to show/hide the code
    df_print: paged
---
  
Script to analyse necklace output using limma. 

Input requited: 

* relST_genomeST.psl file from necklace

* block.counts file from necklace

* X_SraRunTable.txt file with information about the samples 

#Prepare workspace 

Load libraries
```{r message= FALSE, warning=FALSE}
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("tidyr")
#devtools::install_github("sjmgarnier/viridis")
#install.packages("VennDiagram")
#install.packages("rtracklayer")

#source("https://bioconductor.org/biocLite.R")
#biocLite("limma")
#biocLite("edgeR")
#biocLite("DEXSeq")
#biocLite("GenomicRanges")
#biocLite("Rsubread")



library(limma)
library(edgeR)
library(ggplot2)
library(dplyr)
library(plyr)
library(tidyr)
library(DEXSeq)
library(GenomicRanges)
library(rtracklayer)
library(Rsubread)
library(viridis) #colour blind friendly palette for plots
library(VennDiagram)
library(rtracklayer)
library(dplyr)
library(data.table)
library(gplots)
library(UpSetR)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Differential Splicing: Analysis of necklace output of _Apis mellifera_ 
##Pre-processing
###Annotation information

Read in the gtf file 
```{r results="hide", rows.print=5}
ame_gtf <- import.gff3("/Users/afarre/Unimelb/Doublesex/Amellifera/GCF_000002195.4_Amel_4.5_genomic.gff")
ame_gtf$dbGeneIDlist <- paste(ame_gtf$Dbxref, sep=" ", collapse = ",")
as.data.frame(ame_gtf)

```

Keep only gene names and IDs
```{r}
ame_metaData <- as.data.frame(mcols(ame_gtf))

ame_metaData$geneName <- sub("(.*), transcript variant .*", "\\1", ame_metaData$product)

ame_metaDataRNA <- ame_metaData %>%
  dplyr::filter(grepl("rna*", ID)) %>% #keep only rows that match rna* in the ID column
  dplyr::select(type, ID, Name, gene, geneName) %>% #keep only certain columns (others have NA and uninteresting stuff)
  dplyr::group_by(gene, geneName) %>% #group by gene and geneName 
  dplyr::filter(row_number(ID) == 1) %>% #keep only one entry per gene
  dplyr::select(gene, geneName) 

ameMatches <- ame_metaData %>%
  dplyr::filter(grepl("gene*", ID)) %>% #keep only rows that match gene* in the ID column
  dplyr::select(type, ID, gene, dbGeneIDlist) %>% #keep only certain columns (others have NA and uninteresting stuff)
  merge(ame_metaDataRNA, by="gene") %>%
  filter(type!="pseudogene")

ameMatches$dbGeneID <- sub("BEEBASE:(GB\\d*).*", "\\1", ameMatches$dbGeneIDlist)
```

#Analysis of necklace output of _Apis mellifera_ 
##Pre-processing
###Read in the count data
Read in the data (count matrix)

Subset data for SRP046740
```{r  rows.print=5}
counts <- read.csv("/Users/afarre/Unimelb/Doublesex/Bterrestris/necklace/block.counts", sep = "\t", skip = 1, stringsAsFactors = F)
names(counts) <- gsub(".*(ERR.*)\\.bam","\\1", names(counts))

#analyse projects separately
#subset for project SRP046740

counts
```

Convert the counts data frame into a matrix (i.e exclude first 6 columns: gene, chr, etc.)

And create a shorter version of the matrix counts for testing purposes.

```{r rows.print=5}
short_counts <- head(counts)
short_counts_matrix <- short_counts[1:6,7:length(short_counts)]
counts_matrix <- counts[,7:length(counts)]
counts_matrix
dim(counts_matrix)
```

###Create the design matrix

For that we need sample information

Read in the table with all the information of the SRA samples and create "caste" , "sex" and "study" columns.

```{r rows.print=5}
sample_id<-read.table("/Users/afarre/Unimelb/Doublesex/Bterrestris/data_info/Bte_SRR_Acc_List.txt", col.names = "sampleID", stringsAsFactors = F )

info_samples<-read.table("/Users/afarre/Unimelb/Doublesex/Bterrestris/data_info/Bte_SraRunTable.txt", sep = "\t", header = T,  stringsAsFactors = F )

#project SRP046740

info_samples$casteName <- "caste"
info_samples[grep( "MQ*", info_samples$Library_Name, perl=T),]$casteName <- "queen"
info_samples[grep( "MA*", info_samples$Library_Name, perl=T),]$casteName <- "male"
info_samples[grep( "ML.*", info_samples$Library_Name, perl=T),]$casteName <- "male"
info_samples[grep( "MP.*", info_samples$Library_Name, perl=T),]$casteName <- "male"
info_samples[grep( "QL.*", info_samples$Library_Name, perl=T),]$casteName <- "queen"
info_samples[grep( "QP.*", info_samples$Library_Name, perl=T),]$casteName <- "queen"
info_samples[grep( "VQ.*", info_samples$Library_Name, perl=T),]$casteName <- "queen"
info_samples[grep( "WAr.*", info_samples$Library_Name, perl=T),]$casteName <- "worker"
info_samples[grep( "WAu.*", info_samples$Library_Name, perl=T),]$casteName <- "worker"
info_samples[grep( "WL.*", info_samples$Library_Name, perl=T),]$casteName <- "worker"
info_samples[grep( "WP.*", info_samples$Library_Name, perl=T),]$casteName <- "worker"

info_samples$sexName <- "sex"
info_samples[grep( "MQ.*", info_samples$Library_Name, perl=T),]$sexName <- "female"
info_samples[grep( "MA.*", info_samples$Library_Name, perl=T),]$sexName <- "male"
info_samples[grep( "ML.*", info_samples$Library_Name, perl=T),]$sexName <- "male"
info_samples[grep( "MP.*", info_samples$Library_Name, perl=T),]$sexName <- "male"
info_samples[grep( "QL.*", info_samples$Library_Name, perl=T),]$sexName <- "female"
info_samples[grep( "QP.*", info_samples$Library_Name, perl=T),]$sexName <- "female"
info_samples[grep( "VQ.*", info_samples$Library_Name, perl=T),]$sexName <- "female"
info_samples[grep( "WAr.*", info_samples$Library_Name, perl=T),]$sexName <- "female"
info_samples[grep( "WAu.*", info_samples$Library_Name, perl=T),]$sexName <- "female"
info_samples[grep( "WL.*", info_samples$Library_Name, perl=T),]$sexName <- "female"
info_samples[grep( "WP.*", info_samples$Library_Name, perl=T),]$sexName <- "female"

info_samples$stageName <- "stageNameName"
info_samples[grep( "MQ.*", info_samples$Library_Name, perl=T),]$stageName <- "adult"
info_samples[grep( "MA.*", info_samples$Library_Name, perl=T),]$stageName <- "adult"
info_samples[grep( "ML.*", info_samples$Library_Name, perl=T),]$stageName <- "larva"
info_samples[grep( "MP.*", info_samples$Library_Name, perl=T),]$stageName <- "pupa"
info_samples[grep( "QL.*", info_samples$Library_Name, perl=T),]$stageName <- "larva"
info_samples[grep( "QP.*", info_samples$Library_Name, perl=T),]$stageName <- "pupa"
info_samples[grep( "VQ.*", info_samples$Library_Name, perl=T),]$stageName <- "adult"
info_samples[grep( "WAr.*", info_samples$Library_Name, perl=T),]$stageName <- "adult"
info_samples[grep( "WAu.*", info_samples$Library_Name, perl=T),]$stageName <- "adult"
info_samples[grep( "WL.*", info_samples$Library_Name, perl=T),]$stageName <- "larva"
info_samples[grep( "WP.*", info_samples$Library_Name, perl=T),]$stageName <- "pupa"


dplyr::select(info_samples,Library_Name, casteName, sexName, stageName)
```

Create a design matrix

First asign numbers to each level of each factor 

Then specify the design through model.matrix()

```{r rows.print=10}
#Get the sample names per caste and age
#project SRP046740

info_samples$caste <- ""
info_samples[grep( "worker", info_samples$casteName, perl=T),]$caste <- "2"
info_samples[grep( "queen", info_samples$casteName, perl=T),]$caste <- "1"
info_samples[grep( "male", info_samples$casteName, perl=T),]$caste <- "3"

info_samples$sex <- "sex"
info_samples[grep( "queen", info_samples$casteName, perl=T),]$sex <- "1"
info_samples[grep( "male", info_samples$casteName, perl=T),]$sex <- "2"

info_samples$stage <- "stage"
info_samples[grep( "adult", info_samples$stageName, perl=T),]$stage <- "1"
info_samples[grep( "larva", info_samples$stageName, perl=T),]$stage <- "2"
info_samples[grep( "pupa", info_samples$stageName, perl=T),]$stage <- "3"


sample_table <- dplyr::select(info_samples, Run, Library_Name, caste, sex, stage, casteName, sexName, stageName, Colony)
# sample_table_SRP068487 <- dplyr::select(info_samples_SRP068487, age, Run_s, caste, sex, study, ageName, casteName, sexName, studyName)

# data_info <- rbind(sample_table, sample_table_SRP068487)
data_info <- sample_table

row.names(data_info) <- data_info$Run

data_info$caste <- as.factor(data_info$caste)
data_info$stage <- as.factor(data_info$stage)
data_info$Colony <- as.factor(data_info$Colony)

# design <- model.matrix(~ 0 + data_info$caste + data_info$age + data_info$study)

design <- model.matrix(~ 0 + data_info$caste + data_info$stage + data_info$Colony)
row.names(design) <- data_info$Run

#order matrix by row names and change column names to match condition (i.e queen, worker, etc)
design <- design[order(rownames(design)),]
# colnames(design)<- c("queen", "worker", "male", "fourDay", "prepupa", "SRP046740")

colnames(design)<- c("queen", "worker", "male", "larva", "pupa",  "col8", "col9", "col11")
as.data.frame(design)
```



###Create the dge object

Create the dge object with the conts matrix and the gene id information

Counts is a matrix of exon-level counts, and GeneID identities which gene each exon belongs

```{r results="hold"}
dge <- DGEList(counts=counts_matrix, genes=counts$Geneid)
dge$genes$GeneID <- counts$Geneid
dge
```

Include caste and sex info into the DGE object.

```{r results="hide", rows.print=10}


dge$samples$caste <- ""

dge$samples[as.data.frame(design)$queen==1,]$caste <- "1"
dge$samples[as.data.frame(design)$worker==1,]$caste <- "2"
dge$samples[as.data.frame(design)$male==1,]$caste <- "3"

dge$samples$sex <- ""
dge$samples[as.data.frame(design)$queen==1,]$sex <- "1"
dge$samples[as.data.frame(design)$worker==1,]$sex <- "1"
dge$samples[as.data.frame(design)$male==1,]$sex <- "2"

dge$samples$colony <- "3"
dge$samples[as.data.frame(design)$larva==1,]$sex <- "1"
dge$samples[as.data.frame(design)$pupa==1,]$sex <- "2"

dge$samples$sex <- "7"
dge$samples[as.data.frame(design)$col8==1,]$sex <- "8"
dge$samples[as.data.frame(design)$col9==1,]$sex <- "9"
dge$samples[as.data.frame(design)$col11==1,]$sex <- "11"

dge$samples
```

###Calculate normalized library sizes, generate basic descriptive plots

```{r}
dge <- calcNormFactors(dge)
```

Some barplots about mapping stats.

"Raw" library sizes:

```{r}
lib.size <- data.frame(lib.size=sort(colSums(dge$counts)), sample=names(sort(colSums(dge$counts))), stringsAsFactors = F)

lib.size$caste <- data_info[lib.size$sample,]$caste

lib.size$caste <- gsub(pattern = "1", replacement = "Queen", x = lib.size$caste)
lib.size$caste <- gsub(pattern = "2", replacement = "Worker", x = lib.size$caste)
lib.size$caste <- gsub(pattern = "3", replacement = "Male", x = lib.size$caste)

ggplot(lib.size, aes(sample, lib.size, fill=caste))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

```


Normalised library sizes

```{r}
norm.lib.size <- dge$samples
norm.lib.size$sample <- rownames(norm.lib.size) 
norm.lib.size$norm.lib.size <- norm.lib.size$lib.size * norm.lib.size$norm.factors
norm.lib.size <- norm.lib.size[order(norm.lib.size$norm.lib.size),]

norm.lib.size$caste <- gsub(pattern = "1", replacement = "Queen", x = norm.lib.size$caste)
norm.lib.size$caste <- gsub(pattern = "2", replacement = "Worker", x = norm.lib.size$caste)
norm.lib.size$caste <- gsub(pattern = "3", replacement = "Male", x = norm.lib.size$caste)

ggplot(norm.lib.size, aes(sample, norm.lib.size, fill=caste))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

```

####Number of expressed genes per sample

Plot the number of genes expressed in each sample

```{r}
ex_gene_sample <- apply(counts_matrix, 2, function(x){ sum(x!=0) })
ex_gene_sample <- as.data.frame(ex_gene_sample, stringsAsFactors = F)
names(ex_gene_sample) <- "Num_expressed_genes"
ex_gene_sample$sample <- rownames(ex_gene_sample)

ex_gene_sample$caste <- data_info[ex_gene_sample$sample,]$caste

ex_gene_sample$caste <- gsub(pattern = "1", replacement = "Queen", x = ex_gene_sample$caste)
ex_gene_sample$caste <- gsub(pattern = "2", replacement = "Worker", x = ex_gene_sample$caste)
ex_gene_sample$caste <- gsub(pattern = "3", replacement = "Male", x = ex_gene_sample$caste)

ggplot(ex_gene_sample, aes(sample, Num_expressed_genes, fill=caste ))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

```

####Number of expressed genes in % samples

For each gene calculate the % of samples in which it is expressed. 

Calculate how many genes are expressed in each % interval of samples.

```{r}

genes_percent_samples <- as.data.frame(apply(counts_matrix, 1, function(x){ sum(x!=0)/dim(counts_matrix)[2]*100 }))
names(genes_percent_samples) <- "Percent_samples"
genes_percent_samples$gene <- counts$Geneid


ggplot(genes_percent_samples, aes(Percent_samples) )+
  geom_histogram(stat="bin", binwidth=5, aes(fill=..count..))+
  scale_fill_viridis() +
  theme_bw()

```


###Calculate cpm, Normalize and filter

Different filtering either by 

* counts>10 

* cpm>1 in n=3

* cpm>1 in n=(n caste)/2

```{r}
#A <- rowSums(dge$counts)
#dge <- dge[A>10, , keep.lib.sizes=FALSE]

#isexpr <- rowSums(cpm(dge) > 1) >=3
#dge <- dge[isexpr,,keep.lib.sizes=FALSE]

#dge <- calcNormFactors(dge)
```


Calculate log2 CPM
```{r}
cpmNorm <- cpm(dge, normalized.lib.sizes=TRUE, log=TRUE, prior.count=0.25) 
plot.new()
plotDensities(cpmNorm, group = norm.lib.size[colnames(cpmNorm),]$caste) 
abline(v=0)
```

Filter:
  
if n caste <3 then cpm>1 in n=(n caste)

if n caste >3 then cpm>1 in n=(n caste)/2

```{r}
cpm.dge <- dge
cpm.dge$counts <- cpmNorm

#create a normalization factor 
#if #samples < 3 then norm.factor=1
#if #samples > 3 then norm.factor=0.5

norm.samples <- data.frame(caste=levels(as.factor(cpm.dge$samples$caste)), stringsAsFactors = F)
norm.samples$norm <- ""
isexpr <- unlist(lapply(levels(as.factor(cpm.dge$samples$caste)), function(x){
  if(sum(cpm.dge$samples$caste==x)<3){
    1
  }
  else{
    0.5
  }}))
norm.samples$norm <- isexpr

#the normalization factor is used in
# if n caste <3 then cpm>1 in n=(n caste)
# if n caste >3 then cpm>1 in n=(n caste)/2

#which means that we * by the normalization factor 

#cpm>1
dge.filt1 <- dge[
  rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="1",])] >1) >= (sum(cpm.dge$samples$caste=="1")*norm.samples$norm[norm.samples$caste=="1"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="2",])] >1) >= (sum(cpm.dge$samples$caste=="2")*norm.samples$norm[norm.samples$caste=="2"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="3",])] >1) >= (sum(cpm.dge$samples$caste=="3")*norm.samples$norm[norm.samples$caste=="3"])
  ,,keep.lib.sizes=FALSE]

cat(dim(dge.filt1)[1], "exons kept in",dim(dge.filt1)[2],"samples if the filter is set to cmp>1\n")

#cmp>0
dge.filt0 <- dge[
  rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="1",])] >0) >= (sum(cpm.dge$samples$caste=="1")*norm.samples$norm[norm.samples$caste=="1"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="2",])] >0) >= (sum(cpm.dge$samples$caste=="2")*norm.samples$norm[norm.samples$caste=="2"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="3",])] >0) >= (sum(cpm.dge$samples$caste=="3")*norm.samples$norm[norm.samples$caste=="3"])
  ,,keep.lib.sizes=FALSE]

cat(dim(dge.filt0)[1], "exons kept in",dim(dge.filt0)[2],"samples if the filter is set to cmp>0")

```

Recalculate lib size:
  
```{r}
# dge.filt1 <- calcNormFactors(dge.filt1)
dge.filt0 <- calcNormFactors(dge.filt0)
```

##Voom transformation
Apply the voom transformation and fit a linear model
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"No normalization",cex=2,font=2 )
v <- voom(dge.filt0, design, plot=TRUE)

fit <- lmFit(v, design)

v$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v$targets$caste)
v$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v$targets$caste)
v$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v$targets$caste)

contrast.matrix.caste <- makeContrasts(queen-worker, queen-male, worker-male, levels=design)
contrast.matrix.sex <- makeContrasts(male-(queen+worker)/2, levels=design)

fit.caste <- contrasts.fit(fit, contrast.matrix.caste)
fit.caste <- eBayes(fit.caste)

fit.sex <- contrasts.fit(fit, contrast.matrix.sex)
fit.sex <- eBayes(fit.sex)

plot.new()
text(0.5,0.5,"Cyclicloess normalization",cex=2,font=2 )
v.loess <- voom(dge.filt0, design, normalize.method="cyclicloess", plot=TRUE)

fit.loess <- lmFit(v.loess, design)

v.loess$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v.loess$targets$caste)
v.loess$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v.loess$targets$caste)
v.loess$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v.loess$targets$caste)

fit.caste.loess <- contrasts.fit(fit.loess, contrast.matrix.caste)
fit.caste.loess <- eBayes(fit.caste.loess)

fit.sex.loess <- contrasts.fit(fit.loess, contrast.matrix.sex)
fit.sex.loess <- eBayes(fit.sex.loess)

plot.new()
text(0.5,0.5,"Quantile normalization",cex=2,font=2 )
v.quant <- voom(dge.filt0, design, normalize.method="quantile", plot=TRUE)

fit.quant <- lmFit(v.quant, design)

v.quant$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v.quant$targets$caste)
v.quant$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v.quant$targets$caste)
v.quant$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v.quant$targets$caste)

fit.caste.quant <- contrasts.fit(fit.quant, contrast.matrix.caste)
fit.caste.quant <- eBayes(fit.caste.quant)

fit.sex.quant <- contrasts.fit(fit.quant, contrast.matrix.sex)
fit.sex.quant <- eBayes(fit.sex.quant)

```

Densities when using lmFit(v, design)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"No normalization",cex=2,font=2 )
plotDensities(v, group = v$targets$caste)

plot.new()
text(0.5,0.5,"Cyclicloess normalization",cex=2,font=2 )
plotDensities(v.loess, group = v$targets$caste)

plot.new()
text(0.5,0.5,"Quantile normalization",cex=2,font=2 )
plotDensities(v.quant, group = v$targets$caste)



```

Densities when using contrasts on caste

contrasts.fit(fit, contrast.matrix.caste)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste",cex=2,font=2 )
plotDensities(fit.caste)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste Cyclicloess norm.",cex=2,font=2 )
plotDensities(fit.caste.loess)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste quantile norm.",cex=2,font=2 )
plotDensities(fit.caste.quant)
```

Densities when using contrasts on sex

contrasts.fit(fit, contrast.sex)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex",cex=2,font=2 )

plotDensities(fit.sex)
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex Cyclicloess norm.",cex=2,font=2 )
plotDensities(fit.sex.loess)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex quantile norm.",cex=2,font=2 )
plotDensities(fit.sex.quant)
```

```{r}
as.data.frame(contrast.matrix.caste)
```


```{r}
# par(mar=c(1,1,1,1))
# layout( matrix(c(1,2), byrow = TRUE), heights = c(1,3))

# plot.new()
# text(0.5,0.5,"MDS Cyclicloess norm.",cex=2,font=2 )
# plotMDS(v.loess,pch = v$targets$study, col=as.numeric(as.factor(v$targets$caste)))
# plotMDS(v.loess,pch = as.numeric(v$targets$age), col=as.numeric(as.factor(v$targets$caste)))


plot.new()
#text(0.5,0.5,"MDS no norm.",cex=2,font=2 )
# plotMDS(v,pch =  as.numeric(v$targets$age), col=as.numeric(as.factor(v$targets$caste)))
# legend("bottomleft", legend=c(levels(as.factor(v$targets$caste)), "prepupa","2-days","4-days"), text.col =c(1,2,3,1,1,1), pch=c(16,16,16,3,1,2))
plotMDS(v,pch =  16, col=as.numeric(as.factor(v$targets$caste)))
legend("bottomleft", legend=c(levels(as.factor(v$targets$caste))), text.col =c(1,2,3), pch=16)

# plot.new()
# text(0.5,0.5,"MDS quantile norm.",cex=2,font=2 )
# plotMDS(v.quant,pch = 16, col=as.numeric(as.factor(v$targets$caste)))
```

##Differential splicing
###Differentially spliced exons by caste

Now we can test for differential splicing associated with any coeficient in the linear model. 

First run the diffSplice function.
```{r results="hold"}
#ex <- diffSplice(fit, geneid="GeneID")
ex.caste <- diffSplice(fit.caste, geneid="GeneID")

ex.caste.loess <- diffSplice(fit.caste.loess, geneid="GeneID")

ex.caste.quant <- diffSplice(fit.caste.quant, geneid="GeneID")
```

Will find genes that show evidence of differential splicing associated with the first (Queen-Worker) coefficient in the linear model. 

The output is similar that from the limma topTable function. 

```{r rows.print=10}
#More detail can be obtained by topSplice

#First define the number of differentially expressed exons to show
numDiff.caste <- 200000
#then use the function topSplice with the coeficient (comparision) that we're interested in
diff.caste <- topSplice(ex.caste.loess, coef=1, test="simes", number=numDiff.caste,  FDR = 0.05)

diff.caste$ID <- sub("^(gene\\d*):.*", "\\1", diff.caste$genes)


#Find the names of the genes is Apis mellifera 
diffMatches.caste <- diff.caste %>% 
  dplyr::select(ID, P.Value, FDR, NExons) %>%
  merge(ameMatches, by="ID")


in_genes_ds <- diffMatches.caste[grep(".*transformer.*", diffMatches.caste$geneName, ignore.case = TRUE),]
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*vitellogenin.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*bric.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*juvenile hormone.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*insulin.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*royal.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*forkhead.*O-like.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*yellow.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*doublesex.*", diffMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.caste[grep(".*fruitless.*", diffMatches.caste$geneName, ignore.case = TRUE),])
dplyr::select(in_genes_ds, geneName, P.Value, FDR, NExons)

write.table(diffMatches.caste,"/Users/afarre/Unimelb/Doublesex/R/dsCaste_bte.txt", col.names = TRUE)


```

```{r}
fru <- diffMatches.caste[grep( ameMatches[grep(".*fruitless.*", ameMatches$geneName, ignore.case = TRUE),]$gene, diffMatches.caste$gene),]$ID
fru <- ex.caste$genes$GeneID[grep("fru", ex.caste$genes$GeneID)][1]
plotSplice(ex.caste, coef = 1, geneid = fru)
```

###Differential splicing of _doublesex_ by caste

Which will show individual exons that are enriched or depleted relative to other exons in the same gene. 

To display the pattern of exons in the top genes.

Plot doublesex (identified by blast beforehand)  "gene4131:MSTRG.4244"

```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2,3,4,5,6),ncol=1), heights = rep(c(1,3), 3))
#layout.show(n = 6)
plot.new()
text(0.5,0.5,"Queen-Worker",cex=1.5,font=2 )
plotSplice(ex.caste.loess, coef = 1, geneid = "gene4131:MSTRG.4244")

plot.new()
text(0.5,0.5,"Queen-Male",cex=1.5,font=2)
plotSplice(ex.caste.loess, coef = 2, geneid = "gene4131:MSTRG.4244")

plot.new()
text(0.5,0.5,"Worker-Male",cex=1.5,font=2)
plotSplice(ex.caste.loess, coef = 3, geneid = "gene4131:MSTRG.4244")

#dev.off()
```

P values of the exons per contrast

```{r}

#diff$genes=="gene13749:MSTRG.15287"
pvalues_dsx <- ex.caste.loess$p.value[ex.caste.loess$gene.firstexon["gene4131:MSTRG.4244"]:ex.caste.loess$gene.lastexon["gene4131:MSTRG.4244"],]
#rownames(pvalues_dsx) <- c("exon1","exon2","exon3","exon4","exon5","exon6","exon7")
pvalues_dsx
```


###Differential splicing of _feminizer_ by caste

<!-- Which will show individual exons that are enriched or depleted relative to other exons in the same gene.  -->

<!-- To display the pattern of exons in the top genes. -->

<!-- Plot doublesex (identified by blast beforehand)  "gene3145:MSTRG.3247" -->

<!-- ```{r} -->
<!-- par(mar=c(1,1,1,1)) -->
<!-- layout( matrix(c(1,2,3,4,5,6),ncol=1), heights = rep(c(1,3), 3)) -->
<!-- #layout.show(n = 6) -->
<!-- plot.new() -->
<!-- text(0.5,0.5,"Queen-Worker",cex=2,font=2 ) -->
<!-- plotSplice(ex.caste.loess, coef = 1, geneid = "gene3145:MSTRG.3247") -->

<!-- plot.new() -->
<!-- text(0.5,0.5,"Queen-Male",cex=2,font=2) -->
<!-- plotSplice(ex.caste.loess, coef = 2, geneid = "gene3145:MSTRG.3247") -->

<!-- plot.new() -->
<!-- text(0.5,0.5,"Worker-Male",cex=2,font=2) -->
<!-- plotSplice(ex.caste.loess, coef = 3, geneid = "gene3145:MSTRG.3247") -->
<!-- ``` -->

<!-- P values of the exons per contrast -->

<!-- ```{r} -->
<!-- #diff$genes=="gene3145:MSTRG.3247" -->
<!-- pvalues_dsx <- ex.caste.loess$p.value[ex.caste.loess$gene.firstexon["gene3145:MSTRG.3247"]:ex.caste.loess$gene.lastexon["gene3145:MSTRG.3247"],] -->
<!-- #rownames(pvalues_dsx) <- c("exon1","exon2","exon3","exon4","exon5","exon6","exon7") -->
<!-- pvalues_dsx -->
<!-- ``` -->
###Differentially spliced exons by sex

Now we can test for differential splicing associated with any coeficient in the linear model. 

First run the diffSplice function.
```{r results="hold"}
#ex <- diffSplice(fit, geneid="GeneID")
ex.sex <- diffSplice(fit.sex.loess, geneid="GeneID")
```

Will find genes that show evidence of differential splicing associated with the second coefficient in the linear model. 

The output is similar that from the limma topTable function. 

```{r rows.print=10}
#More detail can be obtained by topSplice

#First define the number of differentially expressed exons to show
numDiff.sex <- 200000
#then use the function topSplice with the coeficient (comparision) that we're interested in
diff.sex <- topSplice(ex.sex, coef=1, test="simes", number=numDiff.sex,  FDR = 0.05)

diff.sex$ID <- sub("^(gene\\d*):.*", "\\1", diff.sex$genes)


#Find the names of the genes is Apis mellifera 
diffMatches.sex <- diff.sex %>% 
  dplyr::select(ID, P.Value, FDR, NExons) %>%
  merge(ameMatches, by="ID")


in_genes_ds <- diffMatches.sex[grep(".*transformer.*", diffMatches.sex$geneName, ignore.case = TRUE),]
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*vitellogenin.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*bric.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*juvenile hormone.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*insulin.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*royal.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*forkhead.*O-like.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*yellow.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*doublesex.*", diffMatches.sex$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, diffMatches.sex[grep(".*fruitless.*", diffMatches.sex$geneName, ignore.case = TRUE),])
dplyr::select(in_genes_ds, geneName, P.Value, FDR, NExons)

write.table(diffMatches.sex,"/Users/afarre/Unimelb/Doublesex/R/dsSex_bte.txt", col.names = TRUE)


```

###Differential splicing of _doublesex_ by sex

Which will show individual exons that are enriched or depleted relative to other exons in the same gene. 

To display the pattern of exons in the top genes.

Plot doublesex (identified by blast beforehand)  "gene4131:MSTRG.4244"

```{r}


par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))
#layout.show(n = 2)
plot.new()
text(0.5,0.5,"Male-Female",cex=2,font=2 )
plotSplice(ex.sex, coef = 1, geneid = "gene4131:MSTRG.4244")

#dev.off()

```

P values of the exons per contrast

```{r}
diff.sex[diff.sex=="gene4131:MSTRG.4244",]
#diff$genes=="gene13749:MSTRG.15287"
pvalues_dsx.sex <- ex.sex$p.value[ex.sex$gene.firstexon["gene4131:MSTRG.4244"]:ex.sex$gene.lastexon["gene4131:MSTRG.4244"],]
#names(pvalues_dsx.sex) <- c("exon1","exon2","exon3","exon4","exon5","exon6","exon7")
pvalues_dsx.sex
```

###Intersect Sex and Caste

```{r}
ds_sex_genes <- data.frame(gene=diffMatches.sex$geneName, contrast=rep("sex", length(diffMatches.sex$geneName)))
ds_caste_genes <- data.frame(gene=diffMatches.caste$geneName, contrast=rep("caste", length(diffMatches.caste$geneName)))
ds_matchedgenes <- rbind(ds_caste_genes, ds_sex_genes)
ds_matchedgenes <- unique(ds_matchedgenes)

grid.newpage()
venn.plot <- draw.pairwise.venn(area1 = length(ds_matchedgenes[ds_matchedgenes$contrast=="sex",]$gene), 
                   area2 = length(ds_matchedgenes[ds_matchedgenes$contrast=="caste",]$gene), 
                   cross.area = length(intersect(ds_matchedgenes[ds_matchedgenes$contrast=="sex",]$gene, ds_matchedgenes[ds_matchedgenes$contrast=="caste",]$gene)), 
                   category = c("DS sex", "DS caste"))

length(ds_matchedgenes[ds_matchedgenes$contrast=="sex",]$gene)
length(ds_matchedgenes[ds_matchedgenes$contrast=="caste",]$gene)

# write.table(x = as.data.frame(unlist(Ame_diffMatches.sex)), file = "~/Unimelb/Doublesex/R/Ame_diffMatches_4days_sex.txt", col.names = TRUE, row.names = TRUE)
# write.table(x = as.data.frame(unlist(Ame_diffMatches.caste)), file = "~/Unimelb/Doublesex/R/Ame_diffMatches_4days_caste.txt", col.names = TRUE, row.names = TRUE)

```


#Differential expression: Analysis of necklace output of _Apis mellifera_  

##Pre-processing
###Read in the count data
Read in the data (count matrix)

Subset data for SRP046740
```{r  rows.print=5}
counts <- read.csv("/Users/afarre/Unimelb/Doublesex/Bterrestris/necklace/gene.counts", sep = "\t", skip = 1, stringsAsFactors = F)
names(counts) <- gsub(".*(ERR.*)\\.bam","\\1", names(counts))


```

Convert the counts data frame into a matrix (i.e exclude first 6 columns: gene, chr, etc.)

And create a shorter version of the matrix counts for testing purposes.

```{r rows.print=5}
short_counts <- head(counts)
short_counts_matrix <- short_counts[1:6,7:length(short_counts)]
counts_matrix <- counts[,7:length(counts)]
counts_matrix
dim(counts_matrix)
```

###Create the dge object

Create the dge object with the conts matrix and the gene id information

Counts is a matrix of exon-level counts, and GeneID identities which gene each exon belongs

```{r results="hide"}
dge <- DGEList(counts=counts_matrix, genes=counts$Geneid)
dge$genes$GeneID <- counts$Geneid
dge
```

Include caste and sex info into the DGE object.

```{r results="hide", rows.print=10}
dge$samples$caste <- ""
dge$samples[as.data.frame(design)$queen==1,]$caste <- "1"
dge$samples[as.data.frame(design)$worker==1,]$caste <- "2"
dge$samples[as.data.frame(design)$male==1,]$caste <- "3"

dge$samples$sex <- ""
dge$samples[as.data.frame(design)$queen==1,]$sex <- "1"
dge$samples[as.data.frame(design)$worker==1,]$sex <- "1"
dge$samples[as.data.frame(design)$male==1,]$sex <- "2"

dge$samples
```

###Calculate normalized library sizes, generate basic descriptive plots

```{r}
dge <- calcNormFactors(dge)
```

Some barplots about mapping stats.

"Raw" library sizes:

```{r}
lib.size <- data.frame(lib.size=sort(colSums(dge$counts)), sample=names(sort(colSums(dge$counts))), stringsAsFactors = F)

lib.size$caste <- data_info[lib.size$sample,]$caste

lib.size$caste <- gsub(pattern = "1", replacement = "Queen", x = lib.size$caste)
lib.size$caste <- gsub(pattern = "2", replacement = "Worker", x = lib.size$caste)
lib.size$caste <- gsub(pattern = "3", replacement = "Male", x = lib.size$caste)

ggplot(lib.size, aes(sample, lib.size, fill=caste))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

```


Normalised library sizes

```{r}
norm.lib.size <- dge$samples
norm.lib.size$sample <- rownames(norm.lib.size) 
norm.lib.size$norm.lib.size <- norm.lib.size$lib.size * norm.lib.size$norm.factors
norm.lib.size <- norm.lib.size[order(norm.lib.size$norm.lib.size),]

norm.lib.size$caste <- gsub(pattern = "1", replacement = "Queen", x = norm.lib.size$caste)
norm.lib.size$caste <- gsub(pattern = "2", replacement = "Worker", x = norm.lib.size$caste)
norm.lib.size$caste <- gsub(pattern = "3", replacement = "Male", x = norm.lib.size$caste)

ggplot(norm.lib.size, aes(sample, norm.lib.size, fill=caste))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

```

####Number of expressed genes per sample

Plot the number of genes expressed in each sample

```{r}
ex_gene_sample <- apply(counts_matrix, 2, function(x){ sum(x!=0) })
ex_gene_sample <- as.data.frame(ex_gene_sample, stringsAsFactors = F)
names(ex_gene_sample) <- "Num_expressed_genes"
ex_gene_sample$sample <- rownames(ex_gene_sample)

ex_gene_sample$caste <- data_info[ex_gene_sample$sample,]$caste

ex_gene_sample$caste <- gsub(pattern = "1", replacement = "Queen", x = ex_gene_sample$caste)
ex_gene_sample$caste <- gsub(pattern = "2", replacement = "Worker", x = ex_gene_sample$caste)
ex_gene_sample$caste <- gsub(pattern = "3", replacement = "Male", x = ex_gene_sample$caste)

ggplot(ex_gene_sample, aes(sample, Num_expressed_genes, fill=caste ))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

```

####Number of expressed genes in % samples

For each gene calculate the % of samples in which it is expressed. 

Calculate how many genes are expressed in each % interval of samples.

```{r}

genes_percent_samples <- as.data.frame(apply(counts_matrix, 1, function(x){ sum(x!=0)/dim(counts_matrix)[2]*100 }))
names(genes_percent_samples) <- "Percent_samples"
genes_percent_samples$gene <- counts$Geneid


ggplot(genes_percent_samples, aes(Percent_samples) )+
  geom_histogram(stat="bin", binwidth=5, aes(fill=..count..))+
  scale_fill_viridis() +
  theme_bw()

```


###Calculate cpm, Normalize and filter

Different filtering either by 

* counts>10 

* cpm>1 in n=3

* cpm>1 in n=(n caste)/2

```{r}
#A <- rowSums(dge$counts)
#dge <- dge[A>10, , keep.lib.sizes=FALSE]

#isexpr <- rowSums(cpm(dge) > 1) >=3
#dge <- dge[isexpr,,keep.lib.sizes=FALSE]

#dge <- calcNormFactors(dge)
```


Calculate log2 CPM
```{r}
cpmNorm <- cpm(dge, normalized.lib.sizes=TRUE, log=TRUE, prior.count=0.25) 
plot.new()
plotDensities(cpmNorm, group = norm.lib.size[colnames(cpmNorm),]$caste) 
abline(v=0)
```

Filter:
  
if n caste <3 then cpm>1 in n=(n caste)

if n caste >3 then cpm>1 in n=(n caste)/2

```{r}
cpm.dge <- dge
cpm.dge$counts <- cpmNorm

#create a normalization factor 
#if #samples < 3 then norm.factor=1
#if #samples > 3 then norm.factor=0.5

norm.samples <- data.frame(caste=levels(as.factor(cpm.dge$samples$caste)), stringsAsFactors = F)
norm.samples$norm <- ""
isexpr <- unlist(lapply(levels(as.factor(cpm.dge$samples$caste)), function(x){
  if(sum(cpm.dge$samples$caste==x)<3){
    1
  }
  else{
    0.5
  }}))
norm.samples$norm <- isexpr

#the normalization factor is used in
# if n caste <3 then cpm>1 in n=(n caste)
# if n caste >3 then cpm>1 in n=(n caste)/2

#which means that we * by the normalization factor 

#cpm>2
dge.filt1 <- dge[
  rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="1",])] >1) >= (sum(cpm.dge$samples$caste=="1")*norm.samples$norm[norm.samples$caste=="1"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="2",])] >1) >= (sum(cpm.dge$samples$caste=="2")*norm.samples$norm[norm.samples$caste=="2"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="3",])] >1) >= (sum(cpm.dge$samples$caste=="3")*norm.samples$norm[norm.samples$caste=="3"])
  ,,keep.lib.sizes=FALSE]

cat(dim(dge.filt1)[1], "exons kept in",dim(dge.filt1)[2],"samples if the filter is set to cmp>1\n")

#cmp>0
dge.filt0 <- dge[
  rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="1",])] >0) >= (sum(cpm.dge$samples$caste=="1")*norm.samples$norm[norm.samples$caste=="1"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="2",])] >0) >= (sum(cpm.dge$samples$caste=="2")*norm.samples$norm[norm.samples$caste=="2"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="3",])] >0) >= (sum(cpm.dge$samples$caste=="3")*norm.samples$norm[norm.samples$caste=="3"])
  ,,keep.lib.sizes=FALSE]

cat(dim(dge.filt0)[1], "exons kept in",dim(dge.filt0)[2],"samples if the filter is set to cmp>0")

```

Recalculate lib size:
  
```{r}
# dge.filt1 <- calcNormFactors(dge.filt1)
dge.filt0 <- calcNormFactors(dge.filt0)
```

##Voom transformation
Apply the voom transformation and fit a linear model
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"No normalization",cex=2,font=2 )
v <- voom(dge.filt0, design, plot=TRUE)

fit <- lmFit(v, design)

v$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v$targets$caste)
v$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v$targets$caste)
v$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v$targets$caste)

contrast.matrix.caste <- makeContrasts(queen-worker, queen-male, worker-male, levels=design)
contrast.matrix.sex <- makeContrasts(male-(queen+worker)/2, levels=design)

fit.caste <- contrasts.fit(fit, contrast.matrix.caste)
fit.caste <- eBayes(fit.caste)

fit.sex <- contrasts.fit(fit, contrast.matrix.sex)
fit.sex <- eBayes(fit.sex)

plot.new()
text(0.5,0.5,"Cyclicloess normalization",cex=2,font=2 )
v.loess <- voom(dge.filt0, design, normalize.method="cyclicloess", plot=TRUE)

fit.loess <- lmFit(v.loess, design)

v.loess$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v.loess$targets$caste)
v.loess$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v.loess$targets$caste)
v.loess$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v.loess$targets$caste)

de.fit.caste.loess <- contrasts.fit(fit.loess, contrast.matrix.caste)
de.fit.caste.loess <- eBayes(de.fit.caste.loess)

fit.sex.loess <- contrasts.fit(fit.loess, contrast.matrix.sex)
fit.sex.loess <- eBayes(fit.sex.loess)

plot.new()
text(0.5,0.5,"Quantile normalization",cex=2,font=2 )
v.quant <- voom(dge.filt0, design, normalize.method="quantile", plot=TRUE)

fit.quant <- lmFit(v.quant, design)

v.quant$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v.quant$targets$caste)
v.quant$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v.quant$targets$caste)
v.quant$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v.quant$targets$caste)

fit.caste.quant <- contrasts.fit(fit.quant, contrast.matrix.caste)
fit.caste.quant <- eBayes(fit.caste.quant)

fit.sex.quant <- contrasts.fit(fit.quant, contrast.matrix.sex)
fit.sex.quant <- eBayes(fit.sex.quant)

```

Densities when using lmFit(v, design)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"No normalization",cex=2,font=2 )
plotDensities(v, group = v$targets$caste)

plot.new()
text(0.5,0.5,"Cyclicloess normalization",cex=2,font=2 )
plotDensities(v.loess, group = v$targets$caste)

plot.new()
text(0.5,0.5,"Quantile normalization",cex=2,font=2 )
plotDensities(v.quant, group = v$targets$caste)



```

Densities when using contrasts on caste

contrasts.fit(fit, contrast.matrix.caste)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste",cex=2,font=2 )
plotDensities(fit.caste)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste Cyclicloess norm.",cex=2,font=2 )
plotDensities(de.fit.caste.loess)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste quantile norm.",cex=2,font=2 )
plotDensities(fit.caste.quant)
```

Densities when using contrasts on sex

contrasts.fit(fit, contrast.sex)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex",cex=2,font=2 )

plotDensities(fit.sex)
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex Cyclicloess norm.",cex=2,font=2 )
plotDensities(fit.sex.loess)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex quantile norm.",cex=2,font=2 )
plotDensities(fit.sex.quant)
```

```{r}
as.data.frame(contrast.matrix.caste)
```


```{r}
# par(mar=c(1,1,1,1))
# layout( matrix(c(1,2), byrow = TRUE), heights = c(1,3))

# plot.new()
# text(0.5,0.5,"MDS Cyclicloess norm.",cex=2,font=2 )
# plotMDS(v.loess,pch = v$targets$study, col=as.numeric(as.factor(v$targets$caste)))
# plotMDS(v.loess,pch = as.numeric(v$targets$age), col=as.numeric(as.factor(v$targets$caste)))


plot.new()
#text(0.5,0.5,"MDS no norm.",cex=2,font=2 )
# plotMDS(v,pch =  as.numeric(v$targets$age), col=as.numeric(as.factor(v$targets$caste)))
# legend("bottomleft", legend=c(levels(as.factor(v$targets$caste)), "prepupa","2-days","4-days"), text.col =c(1,2,3,1,1,1), pch=c(16,16,16,3,1,2))
plotMDS(v,pch =  16, col=as.numeric(as.factor(v$targets$caste)))
legend("bottomleft", legend=c(levels(as.factor(v$targets$caste))), text.col =c(1,2,3), pch=16)

# plot.new()
# text(0.5,0.5,"MDS quantile norm.",cex=2,font=2 )
# plotMDS(v.quant,pch = 16, col=as.numeric(as.factor(v$targets$caste)))
```

##Differential expression
###Differential expression by caste

```{r}
#First define the number of differentially expressed exons to show
numDiff.caste <- 200000
#then use the function topSplice with the coeficient (comparision) that we're interested in
de.caste <- topTable(de.fit.caste.loess, coef = 1, number=numDiff.caste,  p.value = 0.05)

length(unique(de.caste$genes))


de.caste$ID <- sub("^(gene\\d*):.*", "\\1", de.caste$genes)


#Find the names of the genes is Apis mellifera 
deMatches.caste <- de.caste %>% 
  dplyr::select(ID, logFC, adj.P.Val) %>%
  merge(ameMatches, by="ID")


in_genes_ds <- deMatches.caste[grep(".*transformer.*", deMatches.caste$geneName, ignore.case = TRUE),]
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*vitellogenin.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*bric.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*juvenile hormone.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*insulin.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*royal.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*forkhead.*O-like.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*yellow.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*doublesex.*", deMatches.caste$geneName, ignore.case = TRUE),])
in_genes_ds <- rbind(in_genes_ds, deMatches.caste[grep(".*fruitless.*", deMatches.caste$geneName, ignore.case = TRUE),])
dplyr::select(in_genes_ds, geneName, ID, logFC, adj.P.Val)

write.table(deMatches.caste,"/Users/afarre/Unimelb/Doublesex/R/deCaste_bte.txt", col.names = TRUE)

ggplot(de.caste, aes(logFC))+
  geom_density()

filter(de.caste, logFC<-1|logFC>1)

```

###Differential expression by sex

```{r}
#First define the number of differentially expressed exons to show
numDiff.sex <- 200000
#then use the function topSplice with the coeficient (comparision) that we're interested in
de.sex <- topTable(fit.sex.loess, coef = 1, number=numDiff.sex,  p.value = 0.05)

de.sex$ID <- sub("^(gene\\d*):.*", "\\1", de.sex$genes)


#Find the names of the genes is Apis mellifera 
deMatches.sex <- de.sex %>% 
  dplyr::select(ID, logFC, adj.P.Val) %>%
  merge(ameMatches, by="ID")


in_genes_de <- deMatches.sex[grep(".*transformer.*", deMatches.sex$geneName, ignore.case = TRUE),]
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*vitellogenin.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*bric.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*juvenile hormone.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*insulin.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*royal.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*forkhead.*O-like.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*yellow.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*doublesex.*", deMatches.sex$geneName, ignore.case = TRUE),])
in_genes_de <- rbind(in_genes_de, deMatches.sex[grep(".*fruitless.*", deMatches.sex$geneName, ignore.case = TRUE),])
dplyr::select(in_genes_de, geneName, ID, logFC, adj.P.Val)

write.table(deMatches.sex,"/Users/afarre/Unimelb/Doublesex/R/deSex_bte.txt", col.names = TRUE)
```

###DE intersect Sex and Caste 

```{r}
de_sex_genes <- data.frame(gene=deMatches.sex$geneName, contrast=rep("sex", length(deMatches.sex$geneName)))
de_caste_genes <- data.frame(gene=deMatches.caste$geneName, contrast=rep("caste", length(deMatches.caste$geneName)))
de_matchedgenes <- rbind(de_caste_genes, de_sex_genes)
de_matchedgenes <- unique(de_matchedgenes)

grid.newpage()
venn.plot <- draw.pairwise.venn(area1 = length(de_matchedgenes[de_matchedgenes$contrast=="sex",]$gene), 
                   area2 = length(de_matchedgenes[de_matchedgenes$contrast=="caste",]$gene), 
                   cross.area = length(intersect(de_matchedgenes[de_matchedgenes$contrast=="sex",]$gene, de_matchedgenes[de_matchedgenes$contrast=="caste",]$gene)), 
                   category = c("DE sex", "DE caste"))
length(unique(deMatches.sex$geneName))
length(unique(deMatches.caste$geneName))

dim(unique(rbind(deMatches.sex, deMatches.sex)))
length(de_matchedgenes[de_matchedgenes$contrast=="sex",]$gene)
length(de_matchedgenes[de_matchedgenes$contrast=="caste",]$gene)
```


Contingency table:

Columns: DE - DS

Rows: Sex - Caste

```{r}
cT <- matrix(c(dim(deMatches.sex)[1], dim(deMatches.caste)[1], dim(diffMatches.sex)[1], dim(diffMatches.caste)[1]), nrow = 2)
cT
```

Test for enrichment: Fisher's exact test

```{r}
fisher.test(cT)

```

Are the genes the same? 

```{r}
deds <- list(de.caste=deMatches.caste$gene, ds.caste=diffMatches.caste$gene,de.sex=deMatches.sex$gene, ds.sex=diffMatches.sex$gene)
venn(deds)
upset(fromList(deds), order.by = "freq")
```

###Doublesex targets

```{r}
#load list of dsx targets
dsxTargets <- read.table("/Users/afarre/Unimelb/Doublesex/R/orthoDB_dsxTargets_dme2ame.txt")

#calculate number of DS genes by caste and by sex
ds.dsx.caste <- unique(diffMatches.caste[diffMatches.caste$dbGeneID %in% dsxTargets$ensembl_id,]$dbGeneID)
ds.dsx.sex <- unique(diffMatches.sex[diffMatches.sex$dbGeneID %in% dsxTargets$ensembl_id,]$dbGeneID)


#differential gene expression by caste

de.dsx.caste <- unique(deMatches.caste[deMatches.caste$dbGeneID %in% dsxTargets$ensembl_id,]$dbGeneID)


#differential gene expression by sex
de.dsx.sex <- unique(deMatches.sex[deMatches.sex$dbGeneID %in% dsxTargets$ensembl_id,]$dbGeneID)

```

Contingency table:

Columns: DE - DS

Rows: Sex - Caste

```{r}
cT <- matrix(c(length(de.dsx.sex), length(de.dsx.caste), length(ds.dsx.sex),length(ds.dsx.caste)), nrow = 2)
cT
```

Test for enrichment: Fisher's exact test

```{r}
cT <- matrix(c(length(diff.caste$genes), length(ds.dsx.caste), 12941, 2533), nrow=2) 
fisher.test(cT)

```

Are the genes the same? 

```{r}
deds <- list(de.dsx.caste=de.dsx.caste, ds.dsx.caste=ds.dsx.caste,de.dsxsex=de.dsx.sex, ds.dsx.sex=ds.dsx.sex)
venn(deds)
upset(fromList(deds), order.by = "freq")
```



Test for enrichment: Fisher's exact test

```{r}
fisher.test(cT)

```

Are the genes the same? 

```{r}
deds <- list(de.caste=de.dsx.caste, ds.caste=ds.dsx.caste,de.sex=de.dsx.sex, ds.sex=ds.dsx.sex)
venn(deds)
upset(fromList(deds), order.by = "freq")
```