---
title: "Intersections_AS"
author: "Anna Orteu"
date: "24/04/2018"
output: 
  html_document:
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 4  # upto 4 depths of headings (specified by #, ##, ### and ####)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: flatly
    code_folding: hide # Add some awesome buttons to show/hide the code
    df_print: paged
---

#Prepare workspace

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message= FALSE, warning=FALSE}
# install.packages("UpSetR")
library(UpSetR)
library(reshape2)
library(ggplot2)
library(plyr)
```

# Intersection plots

## Caste all species

Plot intersections of genes that present alternative splicing between castes (Queens and workers) and that have a match in *Drosophila melanogaster* genome between species.

Species used:

Name used|Species|Family|Subfamily
:--|:--|:--|:--
Ame|*Apis mellifera*|Apidae|
Mph|*Monomorium pharaonis*|Formicidae|Myrmicinae
Wau|*Wasmannia auropunctata*|Formicidae|Myrmicinae
Aec|*Acromyrmex echinatior*|Formicidae|Myrmicinae
Vem|*Vollenhovia emeryi*|Formicidae|Myrmicinae


```{r}
#load the data
Ame_diffMatches.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_caste.txt", header = T)
Ame_diffMatches_2_4days.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_2_4_days_caste.txt", header = T)
Ame_diffMatches_2days.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_2days_caste.txt", header = T)
Ame_diffMatches_4days.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_4days_caste.txt", header = T)
Wau_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Wau_diffMatches_caste.txt", header = T)
Mph_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Mph_diffMatches_DRP002877_caste.txt", header = T)
Aec_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Aec_diffMatches_casteGyneWorker.txt", header = T)
Vem_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Vem_diffMatches_caste.txt", header = T)

Ame_diffMatches.caste.df <- data.frame(species="Ame.prepupa", genes=unlist(Ame_diffMatches.caste))
Ame_diffMatches_2days.caste.df <- data.frame(species="Ame.2days", genes=unlist(Ame_diffMatches_2days.caste))
Ame_diffMatches_4days.caste.df <- data.frame(species="Ame.4days", genes=unlist(Ame_diffMatches_4days.caste))
Wau_diffMatches.caste.df <- data.frame(species="Wau", genes=unlist(Wau_diffMatches.caste))
Mph_diffMatches.caste.df <- data.frame(species="Mph", genes=unlist(Mph_diffMatches.caste))
Aec_diffMatches.caste.df <- data.frame(species="Aec", genes=unlist(Aec_diffMatches.caste))
Vem_diffMatches.caste.df <- data.frame(species="Vem", genes=unlist(Vem_diffMatches.caste))

#create a presence/abscence df
df.caste <- rbind(unique(Ame_diffMatches.caste.df), unique(Ame_diffMatches_2days.caste.df))
df.caste <- rbind(df.caste, unique(Ame_diffMatches_4days.caste.df))
df.caste <- rbind(df.caste, unique(Wau_diffMatches.caste.df))
df.caste <- rbind(df.caste, unique(Mph_diffMatches.caste.df))
df.caste <- rbind(df.caste, unique(Aec_diffMatches.caste.df))
df.caste <- rbind(df.caste, unique(Vem_diffMatches.caste.df))

df.caste <- dcast(df.caste, genes~species, length)

# listInput.caste <- list(Ame.prepupa=Ame_diffMatches.caste$unlist.Ame_diffMatches.caste., 
#                   Ame.2day=Ame_diffMatches_2days.caste$unlist.Ame_diffMatches.caste.,
#                   Ame.4day=Ame_diffMatches_4days.caste$unlist.Ame_diffMatches.caste.,
#                   Wau=Wau_diffMatches.caste$unlist.Wau_diffMatches.caste., 
#                   Aec=Aec_diffMatches.caste$unlist.Aec_diffMatches.sex., 
#                   Mph=Mph_diffMatches.caste$unlist.Mph_diffMatches.DRP002877.caste.,
#                   Vem=Vem_diffMatches.caste$unlist.Vem_diffMatches.caste.)

# upset(fromList(listInput.caste), order.by = "freq", nsets = 7)
# names(df.caste)[1] <- "Name"
# matrix.caste <- df.caste
# rownames(matrix.caste) <- matrix.caste$Name
# matrix.caste <- matrix.caste[-1]

#plot
upset(df.caste, order.by = "freq", nsets = 7)

      # queries = list(list(query = intersects, 
      #                     params = list("Vem"), 
      #                     color = "orange", active = T)))



```

Percentage of genes only found in one of the species and none of the others: 

```{r}
matrix.caste <- df.caste
rownames(matrix.caste) <- matrix.caste$genes
matrix.caste <- matrix.caste[-1]

vem.percent <- data.frame(num.genes=c(dim(matrix.caste[matrix.caste$Vem==1&rowSums(matrix.caste)==1,])[1],
                 dim(matrix.caste[matrix.caste$Vem==1&rowSums(matrix.caste)==2,])[1],
                 dim(matrix.caste[matrix.caste$Vem==1&rowSums(matrix.caste)==3,])[1],
                 dim(matrix.caste[matrix.caste$Vem==1&rowSums(matrix.caste)==4,])[1],
                 dim(matrix.caste[matrix.caste$Vem==1&rowSums(matrix.caste)==5,])[1],
                 dim(matrix.caste[matrix.caste$Vem==1&rowSums(matrix.caste)==6,])[1],
                 dim(matrix.caste[matrix.caste$Vem==1&rowSums(matrix.caste)==7,])[1]),
           species="Vem", species.num=c("1","2","3","4","5","6","7"))

mph.percent <- data.frame(num.genes=c(dim(matrix.caste[matrix.caste$Mph==1&rowSums(matrix.caste)==1,])[1],
                 dim(matrix.caste[matrix.caste$Mph==1&rowSums(matrix.caste)==2,])[1],
                 dim(matrix.caste[matrix.caste$Mph==1&rowSums(matrix.caste)==3,])[1],
                 dim(matrix.caste[matrix.caste$Mph==1&rowSums(matrix.caste)==4,])[1],
                 dim(matrix.caste[matrix.caste$Mph==1&rowSums(matrix.caste)==5,])[1],
                 dim(matrix.caste[matrix.caste$Mph==1&rowSums(matrix.caste)==6,])[1],
                 dim(matrix.caste[matrix.caste$Mph==1&rowSums(matrix.caste)==7,])[1]),
           species="Mph", species.num=c("1","2","3","4","5","6","7"))

wau.percent <- data.frame(num.genes=c(dim(matrix.caste[matrix.caste$Wau==1&rowSums(matrix.caste)==1,])[1],
                 dim(matrix.caste[matrix.caste$Wau==1&rowSums(matrix.caste)==2,])[1],
                 dim(matrix.caste[matrix.caste$Wau==1&rowSums(matrix.caste)==3,])[1],
                 dim(matrix.caste[matrix.caste$Wau==1&rowSums(matrix.caste)==4,])[1],
                 dim(matrix.caste[matrix.caste$Wau==1&rowSums(matrix.caste)==5,])[1],
                 dim(matrix.caste[matrix.caste$Wau==1&rowSums(matrix.caste)==6,])[1],
                 dim(matrix.caste[matrix.caste$Wau==1&rowSums(matrix.caste)==7,])[1]),
           species="Wau", species.num=c("1","2","3","4","5","6","7"))

ame.prepupa.percent <- data.frame(num.genes=c(dim(matrix.caste[matrix.caste$Ame.prepupa==1&rowSums(matrix.caste)==1,])[1],
                 dim(matrix.caste[matrix.caste$Ame.prepupa==1&rowSums(matrix.caste)==2,])[1],
                 dim(matrix.caste[matrix.caste$Ame.prepupa==1&rowSums(matrix.caste)==3,])[1],
                 dim(matrix.caste[matrix.caste$Ame.prepupa==1&rowSums(matrix.caste)==4,])[1],
                 dim(matrix.caste[matrix.caste$Ame.prepupa==1&rowSums(matrix.caste)==5,])[1],
                 dim(matrix.caste[matrix.caste$Ame.prepupa==1&rowSums(matrix.caste)==6,])[1],
                 dim(matrix.caste[matrix.caste$Ame.prepupa==1&rowSums(matrix.caste)==7,])[1]),
           species="Ame.prepupa", species.num=c("1","2","3","4","5","6","7"))

ame2day.percent <- data.frame(num.genes=c(dim(matrix.caste[matrix.caste$Ame.2days==1&rowSums(matrix.caste)==1,])[1],
                 dim(matrix.caste[matrix.caste$Ame.2days==1&rowSums(matrix.caste)==2,])[1],
                 dim(matrix.caste[matrix.caste$Ame.2days==1&rowSums(matrix.caste)==3,])[1],
                 dim(matrix.caste[matrix.caste$Ame.2days==1&rowSums(matrix.caste)==4,])[1],
                 dim(matrix.caste[matrix.caste$Ame.2days==1&rowSums(matrix.caste)==5,])[1],
                 dim(matrix.caste[matrix.caste$Ame.2days==1&rowSums(matrix.caste)==6,])[1],
                 dim(matrix.caste[matrix.caste$Ame.2days==1&rowSums(matrix.caste)==7,])[1]),
           species="Ame.2days", species.num=c("1","2","3","4","5","6","7"))

ame4day.percent <- data.frame(num.genes=c(dim(matrix.caste[matrix.caste$Ame.4days==1&rowSums(matrix.caste)==1,])[1],
                 dim(matrix.caste[matrix.caste$Ame.4days==1&rowSums(matrix.caste)==2,])[1],
                 dim(matrix.caste[matrix.caste$Ame.4days==1&rowSums(matrix.caste)==3,])[1],
                 dim(matrix.caste[matrix.caste$Ame.4days==1&rowSums(matrix.caste)==4,])[1],
                 dim(matrix.caste[matrix.caste$Ame.4days==1&rowSums(matrix.caste)==5,])[1],
                 dim(matrix.caste[matrix.caste$Ame.4days==1&rowSums(matrix.caste)==6,])[1],
                 dim(matrix.caste[matrix.caste$Ame.4days==1&rowSums(matrix.caste)==7,])[1]),
           species="Ame.4days", species.num=c("1","2","3","4","5","6","7"))

aec.percent <- data.frame(num.genes=c(dim(matrix.caste[matrix.caste$Aec==1&rowSums(matrix.caste)==1,])[1],
                 dim(matrix.caste[matrix.caste$Aec==1&rowSums(matrix.caste)==2,])[1],
                 dim(matrix.caste[matrix.caste$Aec==1&rowSums(matrix.caste)==3,])[1],
                 dim(matrix.caste[matrix.caste$Aec==1&rowSums(matrix.caste)==4,])[1],
                 dim(matrix.caste[matrix.caste$Aec==1&rowSums(matrix.caste)==5,])[1],
                 dim(matrix.caste[matrix.caste$Aec==1&rowSums(matrix.caste)==6,])[1],
                 dim(matrix.caste[matrix.caste$Aec==1&rowSums(matrix.caste)==7,])[1]),
           species="Aec", species.num=c("1","2","3","4","5","6","7"))

df.percent <- rbind(vem.percent, wau.percent)
df.percent <- rbind(df.percent, mph.percent)
df.percent <- rbind(df.percent, ame.prepupa.percent)
df.percent <- rbind(df.percent, ame4day.percent)
df.percent <- rbind(df.percent, ame2day.percent)
df.percent <- rbind(df.percent, aec.percent)


ggplot(df.percent, aes(species, num.genes, fill=species.num))+
  geom_bar(stat="identity", position = "fill")+
  ylab("% genes")

ggplot(df.percent, aes(species, num.genes, fill=species.num))+
  geom_bar(stat="identity")+
  ylab("Number of genes")

```



2-day old *Apis mellifera* larvae present only ~300 AS genes between castes. All of them are found in caste comparisons in other species.

Simplify the plot by excluding the 2-day old *Apis mellifera* larvae: 

```{r}
Ame_diffMatches.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_caste.txt", header = T)
Ame_diffMatches_2_4days.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_2_4_days_caste.txt", header = T)
Ame_diffMatches_2days.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_2days_caste.txt", header = T)
Ame_diffMatches_4days.caste <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_4days_caste.txt", header = T)
Wau_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Wau_diffMatches_caste.txt", header = T)
Mph_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Mph_diffMatches_DRP002877_caste.txt", header = T)
Aec_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Aec_diffMatches_casteGyneWorker.txt", header = T)
Vem_diffMatches.caste <-  read.table("~/Unimelb/Doublesex/R/Vem_diffMatches_caste.txt", header = T)

listInput.caste <- list(Ame.prepupa=Ame_diffMatches.caste$unlist.Ame_diffMatches.caste., 
                  Ame.4day=Ame_diffMatches_4days.caste$unlist.Ame_diffMatches.caste.,
                  Wau=Wau_diffMatches.caste$unlist.Wau_diffMatches.caste., 
                  Aec=Aec_diffMatches.caste$unlist.Aec_diffMatches.sex., 
                  Mph=Mph_diffMatches.caste$unlist.Mph_diffMatches.DRP002877.caste.,
                  Vem=Vem_diffMatches.caste$unlist.Vem_diffMatches.caste.)

upset(fromList(listInput.caste), order.by = "freq", nsets = 6)
```

## Caste *Apis mellifera*

Compare the different stages of *Apis mellifera*. Are AS genes between castes shared between stages?

```{r}
listInput.ame <- list(Ame.prepupa=Ame_diffMatches.caste$unlist.Ame_diffMatches.caste., 
                  Ame.2day=Ame_diffMatches_2days.caste$unlist.Ame_diffMatches.caste.,
                  Ame.4day=Ame_diffMatches_4days.caste$unlist.Ame_diffMatches.caste.)

upset(fromList(listInput.ame), order.by = "freq", empty.intersections = T)
```


## Sex all species 

Plot intersections of genes that present alternative splicing between sexes and that have a match in *Drosophila melanogaster* genome between species.

```{r}
Ame_diffMatches.sex <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_sex.txt", header = T)
Ame_diffMatches_2_4days.sex <- read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_2_4_days_sex.txt", header = T)
Ame_diffMatches_2days.sex <-  read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_2days_sex.txt", header = T)
Ame_diffMatches_4days.sex <-  read.table("~/Unimelb/Doublesex/R/Ame_diffMatches_4days_sex.txt", header = T)
Wau_diffMatches.sex <-  read.table("~/Unimelb/Doublesex/R/Wau_diffMatches_sex.txt", header = T)
Vem_diffMatches.sex <-  read.table("~/Unimelb/Doublesex/R/Vem_diffMatches_sex.txt", header = T)



listInput.sex <- list(Ame.prepupa=Ame_diffMatches.sex$unlist.Ame_diffMatches.sex., 
                  Ame.2day=Ame_diffMatches_2days.sex$unlist.Ame_diffMatches.sex.,
                  Ame.4day=Ame_diffMatches_4days.sex$unlist.Ame_diffMatches.sex.,
                  Wau=Wau_diffMatches.sex$unlist.Wau_diffMatches.sex.,
                  Vem=Vem_diffMatches.sex$unlist.Vem_diffMatches.sex.)

upset(fromList(listInput.sex), order.by = "freq")
```

```{r}
Ame_diffMatches.sex.df <- data.frame(species="Ame.prepupa", genes=unlist(Ame_diffMatches.sex))
Ame_diffMatches_2days.sex.df <- data.frame(species="Ame.2days", genes=unlist(Ame_diffMatches_2days.sex))
Ame_diffMatches_4days.sex.df <- data.frame(species="Ame.4days", genes=unlist(Ame_diffMatches_4days.sex))
Wau_diffMatches.sex.df <- data.frame(species="Wau", genes=unlist(Wau_diffMatches.sex))
Vem_diffMatches.sex.df <- data.frame(species="Vem", genes=unlist(Vem_diffMatches.sex))


df.sex <- rbind(unique(Ame_diffMatches.sex.df), unique(Ame_diffMatches_2days.sex.df))
df.sex <- rbind(df.sex, unique(Ame_diffMatches_4days.sex.df))
df.sex <- rbind(df.sex, unique(Wau_diffMatches.sex.df))
df.sex <- rbind(df.sex, unique(Vem_diffMatches.sex.df))

df.sex <- dcast(df.sex, genes~species, length)

matrix.sex <- df.sex
rownames(matrix.sex) <- matrix.sex$genes
matrix.sex <- matrix.sex[-1]

vem.percent.sex <- data.frame(num.genes=c(dim(matrix.sex[matrix.sex$Vem==1&rowSums(matrix.sex)==1,])[1],
                 dim(matrix.sex[matrix.sex$Vem==1&rowSums(matrix.sex)==2,])[1],
                 dim(matrix.sex[matrix.sex$Vem==1&rowSums(matrix.sex)==3,])[1],
                 dim(matrix.sex[matrix.sex$Vem==1&rowSums(matrix.sex)==4,])[1],
                 dim(matrix.sex[matrix.sex$Vem==1&rowSums(matrix.sex)==5,])[1]),
           species="Vem", species.num=c("1","2","3","4","5"))

wau.percent.sex <- data.frame(num.genes=c(dim(matrix.sex[matrix.sex$Wau==1&rowSums(matrix.sex)==1,])[1],
                 dim(matrix.sex[matrix.sex$Wau==1&rowSums(matrix.sex)==2,])[1],
                 dim(matrix.sex[matrix.sex$Wau==1&rowSums(matrix.sex)==3,])[1],
                 dim(matrix.sex[matrix.sex$Wau==1&rowSums(matrix.sex)==4,])[1],
                 dim(matrix.sex[matrix.sex$Wau==1&rowSums(matrix.sex)==5,])[1]),
           species="Wau", species.num=c("1","2","3","4","5"))

ame.prepupa.percent.sex <- data.frame(num.genes=c(dim(matrix.sex[matrix.sex$Ame.prepupa==1&rowSums(matrix.sex)==1,])[1],
                 dim(matrix.sex[matrix.sex$Ame.prepupa==1&rowSums(matrix.sex)==2,])[1],
                 dim(matrix.sex[matrix.sex$Ame.prepupa==1&rowSums(matrix.sex)==3,])[1],
                 dim(matrix.sex[matrix.sex$Ame.prepupa==1&rowSums(matrix.sex)==4,])[1],
                 dim(matrix.sex[matrix.sex$Ame.prepupa==1&rowSums(matrix.sex)==5,])[1]),
           species="Ame.prepupa", species.num=c("1","2","3","4","5"))

ame2day.percent.sex <- data.frame(num.genes=c(dim(matrix.sex[matrix.sex$Ame.2days==1&rowSums(matrix.sex)==1,])[1],
                 dim(matrix.sex[matrix.sex$Ame.2days==1&rowSums(matrix.sex)==2,])[1],
                 dim(matrix.sex[matrix.sex$Ame.2days==1&rowSums(matrix.sex)==3,])[1],
                 dim(matrix.sex[matrix.sex$Ame.2days==1&rowSums(matrix.sex)==4,])[1],
                 dim(matrix.sex[matrix.sex$Ame.2days==1&rowSums(matrix.sex)==5,])[1]),
           species="Ame.2days", species.num=c("1","2","3","4","5"))

ame4day.percent.sex <- data.frame(num.genes=c(dim(matrix.sex[matrix.sex$Ame.4days==1&rowSums(matrix.sex)==1,])[1],
                 dim(matrix.sex[matrix.sex$Ame.4days==1&rowSums(matrix.sex)==2,])[1],
                 dim(matrix.sex[matrix.sex$Ame.4days==1&rowSums(matrix.sex)==3,])[1],
                 dim(matrix.sex[matrix.sex$Ame.4days==1&rowSums(matrix.sex)==4,])[1],
                 dim(matrix.sex[matrix.sex$Ame.4days==1&rowSums(matrix.sex)==5,])[1]),
           species="Ame.4days", species.num=c("1","2","3","4","5"))

df.percent.sex <- rbind(vem.percent.sex, wau.percent.sex)
df.percent.sex <- rbind(df.percent.sex, ame.prepupa.percent.sex)
df.percent.sex <- rbind(df.percent.sex, ame4day.percent.sex)
df.percent.sex <- rbind(df.percent.sex, ame2day.percent.sex)

ggplot(df.percent.sex, aes(species, num.genes, fill=species.num))+
  geom_bar(stat="identity", position = "fill")+
  ylab("% genes")

ggplot(df.percent.sex, aes(species, num.genes, fill=species.num))+
  geom_bar(stat="identity")+
  ylab("Number of genes")
```


## Sex *Apis mellifera*

Compare sexes in *Apis mellifera*. Are AS genes between sexes shared between stages?

```{r}
listInput.ame.sex <- list(Ame.prepupa=Ame_diffMatches.sex$unlist.Ame_diffMatches.sex., 
                  Ame.2day=Ame_diffMatches_2days.sex$unlist.Ame_diffMatches.sex.,
                  Ame.4day=Ame_diffMatches_4days.sex$unlist.Ame_diffMatches.sex.)

upset(fromList(listInput.ame.sex), order.by = "freq", empty.intersections = T)
```
