---
title: "Vem_necklace_limma"
author: "Anna Orteu"
date: "25/04/2018"
output: 
  html_document:
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 4  # upto 4 depths of headings (specified by #, ##, ### and ####)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: flatly
    code_folding: hide # Add some awesome buttons to show/hide the code
    df_print: paged
---

Script to analyse necklace output using limma. 
Made for non-model insects - find orthology to _Drosophila_ genome. 

Input requited: 

* relST_genomeST.psl file from necklace

* _Drosophila_ gtf file Drosophila_melanogaster.BDGP6.37.gtf

* block.counts file from necklace

* X_SraRunTable.txt file with information about the samples 

#Prepare workspace 

Load libraries
```{r message= FALSE, warning=FALSE}
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("tidyr")
#devtools::install_github("sjmgarnier/viridis")
#install.packages("VennDiagram")
#install.packages("rtracklayer")

#source("https://bioconductor.org/biocLite.R")
#biocLite("limma")
#biocLite("edgeR")
#biocLite("DEXSeq")
#biocLite("GenomicRanges")
#biocLite("Rsubread")



library(limma)
library(edgeR)
library(ggplot2)
library(dplyr)
library(plyr)
library(tidyr)
library(DEXSeq)
library(GenomicRanges)
library(rtracklayer)
library(Rsubread)
library(viridis) #colour blind friendly palette for plots
library(VennDiagram)
library(rtracklayer)
library(dplyr)
library(data.table)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Gene orthology
##The datafiles: psl and gtf

This section of the script uses 2 types of files: 
 
 1. **Psl**: output from blat 
         Blat done with genes assembled with StringTie in a non-model species
         matched to a model species (e.g. _Drosophila melanogaster_)

* **MSTRG** -> StringTie genes in the non-model
   
* **FB** -> FlyBase genes in the model species

 2. **Gtf**: gtf file from the model species 
         
Contains the gene ids from FlyBase FB also the gene names (e.g. dsx)

**Goal:** 

    MSTRG#(non-model) -> FB#(model) -> gene_name(model)


Read in the psl file (output file from blat)
```{r}
dataFile = read.table('/Users/afarre/Unimelb/Doublesex/Vemeryi/necklace/extra_steps_necklace/relST_genomeST.psl', header = FALSE, sep = "\t", skip = 5, stringsAsFactors = F)
```

Psl file has no header. Create vector of column names
```{r}
names_psl <- unlist(strsplit("matches
misMatches
repMatches
nCount
qNumInsert
qBaseInsert
tNumInsert
tBaseInsert
strand
qName
qSize
qStart
qEnd
tName
tSize
tStart
tEnd
blockCount
blockSizes
qStarts
tStarts", "\n"))
```

Change names of the psl object. 

Columns with query (non-model species) and target (model species) gene names:
```{r cols.print=3, rows.print=3}
names(dataFile) <- names_psl
dplyr::select(dataFile[14:18,], qName, tName)
```

Read in the gtf file 
```{r rows.print=5}
dme_gtf <- import("/Users/afarre/Unimelb/Doublesex/Drosophila_melanogaster.BDGP6.37.gtf")
as.data.frame(dme_gtf)
```

##From the genes of the non-model organism to the model's

Using the psl file we get the matches of the StringTie genes to the FlyBase ones

Get all stringtie gene names in a new column
```{r rows.print=5}
dataFile$qmstrg <- sub(".*(MSTRG\\.\\d*).*", "\\1", dataFile$qName)
as.data.frame(unique(dataFile$qmstrg))
```

Get all FlyBase (FB*) gene ids in a list, each entry corresponds to the FB gene match
```{r results="hold"}
listFB <- strsplit(dataFile$tName, ":") 
names(listFB) <- dataFile$qmstrg
listFB <- lapply(listFB, function(x){x[grep("^FBgn\\d*", x)]})

head(listFB)
```

##From FlyBase IDs to gene names in _Drosophila melanogaster_

Using the gtf file 

For each gene, find the gene name using the FB gene id
```{r rows.print=5}
mcols(dme_gtf)$gene_idFB <- sub(".*(FBgn\\d*).*", "\\1", mcols(dme_gtf)$gene_id)
as.data.frame((mcols(dme_gtf)$gene_idFB))
```

remove all entries that contain NA in the "gene_id" column
```{r rows.print=5}
mcols(dme_gtf) <- as.data.frame(na.omit(as.data.table(mcols(dme_gtf)), cols="gene_idFB"))

as.data.frame(mcols(dme_gtf))
```

For each StringTie gene in the Query (non-model) species find the match in the Related (model) species
```{r results="hold"}
matchList <- lapply(listFB, function(listelement){sapply(listelement, function(x){unique(mcols(dme_gtf[mcols(dme_gtf)$gene_idFB==x])$gene_name)})})

head(matchList)
```



#Analysis of necklace output of _Vollenhovia emeryi_ 
##Pre-processing
###Read in the count data
Read in the data (count matrix)
```{r  rows.print=5}
counts <- read.csv("/Users/afarre/Unimelb/Doublesex/Vemeryi/necklace/extra_steps_necklace/block.counts", sep = "\t", skip = 1, stringsAsFactors = F)
names(counts) <- gsub(".*(DRR.*)\\.bam","\\1", names(counts))
counts
```


Exclude the pooled sample, DRR029092 (because is a mix of individuals). 

Convert the counts data frame into a matrix (i.e exclude first 6 columns: gene, chr, etc.)

And create a shorter version of the matrix counts for testing purposes.

```{r rows.print=5}
short_counts <- head(counts)
short_counts_matrix <- short_counts[1:6,7:length(short_counts)]
counts_matrix <- counts[,7:length(counts)]
counts_matrix
dim(counts_matrix)
```


###Create the dge object

Create the dge object with the conts matrix and the gene id information

Counts is a matrix of exon-level counts, and GeneID identities which gene each exon belongs

```{r results="hold"}
dge <- DGEList(counts=counts_matrix, genes=counts$Geneid)
dge$genes$GeneID <- counts$Geneid
dge
```

###Create the design matrix

For that we need sample information

Read in the table with all the information of the SRA samples

Exclude the pooled sample, DRR029092

```{r rows.print=5}
data_info<-read.csv("/Users/afarre/Unimelb/Doublesex/Vemeryi/data_info/Vem_DRP002519_SraRunTable.txt", sep = "\t")
rownames(data_info) <- data_info$Run_s
data_info
```

Create a design matrix

First asign numbers to each level of each factor 

Then specify the design through model.matrix()

```{r rows.print=10}
data_info$caste <- "1"
#data_info[grep("Queen", data_info$sample_title_s),]$caste <- "1"
data_info[grep("Worker", data_info$sample_title_s),]$caste <- "2"
data_info[grep("Male", data_info$sample_title_s),]$caste <- "3"


design <- model.matrix(~ 0 + data_info$caste)
row.names(design) <- data_info$Run_s

#order matrix by row names and change column names to match condition (i.e queen, worker, etc)
design <- design[order(rownames(design)),]
colnames(design)<- c("queen", "worker", "male")

as.data.frame(design)
```

Include caste info into the DGE object.

```{r rows.print=10}
dge$samples$caste <- ""
dge$samples[as.data.frame(design)$queen==1,]$caste <- "1"
dge$samples[as.data.frame(design)$worker==1,]$caste <- "2"
dge$samples[as.data.frame(design)$male==1,]$caste <- "3"

dge$samples
```


###Calculate normalized library sizes, generate basic descriptive plots

```{r}
dge <- calcNormFactors(dge)
```

Some barplots about mapping stats.

"Raw" library sizes:
```{r}
lib.size <- data.frame(lib.size=sort(colSums(dge$counts)), sample=names(sort(colSums(dge$counts))), stringsAsFactors = F)

lib.size$caste <- data_info[lib.size$sample,]$caste

lib.size$caste <- gsub(pattern = "1", replacement = "Queen", x = lib.size$caste)
lib.size$caste <- gsub(pattern = "2", replacement = "Worker", x = lib.size$caste)
lib.size$caste <- gsub(pattern = "3", replacement = "Male", x = lib.size$caste)

ggplot(lib.size, aes(sample, lib.size, fill=caste))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))

```


Normalised library sizes

```{r}
norm.lib.size <- dge$samples
norm.lib.size$sample <- rownames(norm.lib.size) 
norm.lib.size$norm.lib.size <- norm.lib.size$lib.size * norm.lib.size$norm.factors
norm.lib.size <- norm.lib.size[order(norm.lib.size$norm.lib.size),]

norm.lib.size$caste <- gsub(pattern = "1", replacement = "Queen", x = norm.lib.size$caste)
norm.lib.size$caste <- gsub(pattern = "2", replacement = "Worker", x = norm.lib.size$caste)
norm.lib.size$caste <- gsub(pattern = "3", replacement = "Male", x = norm.lib.size$caste)

ggplot(norm.lib.size, aes(sample, norm.lib.size, fill=caste))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))

```

####Number of expressed blocks per sample

Plot the number of blocks expressed in each sample

```{r}
ex_gene_sample <- apply(counts_matrix, 2, function(x){ sum(x!=0) })
ex_gene_sample <- as.data.frame(ex_gene_sample, stringsAsFactors = F)
names(ex_gene_sample) <- "Num_expressed_blocks"
ex_gene_sample$sample <- rownames(ex_gene_sample)

ex_gene_sample$caste <- data_info[ex_gene_sample$sample,]$caste

ex_gene_sample$caste <- gsub(pattern = "1", replacement = "Queen", x = ex_gene_sample$caste)
ex_gene_sample$caste <- gsub(pattern = "2", replacement = "Worker", x = ex_gene_sample$caste)
ex_gene_sample$caste <- gsub(pattern = "3", replacement = "Male", x = ex_gene_sample$caste)

ggplot(ex_gene_sample, aes(sample, Num_expressed_genes, fill=caste ))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))

```

####Number of expressed blocks in % samples

For each block calculate the % of samples in which it is expressed. 

Calculate how many blocks are expressed in each % interval of samples.

```{r}

genes_percent_samples <- as.data.frame(apply(counts_matrix, 1, function(x){ sum(x!=0)/dim(counts_matrix)[2]*100 }))
names(genes_percent_samples) <- "Percent_samples"
genes_percent_samples$gene <- counts$Geneid


ggplot(genes_percent_samples, aes(Percent_samples) )+
  geom_histogram(stat="bin", binwidth=5, aes(fill=..count..))+
  scale_fill_viridis() +
  theme_bw()

```


###Calculate cpm, Normalize and filter

Different filtering either by 

* counts>10 

* cpm>1 in n=3

* cpm>1 in n=(n caste)/2

```{r}
#A <- rowSums(dge$counts)
#dge <- dge[A>10, , keep.lib.sizes=FALSE]

#isexpr <- rowSums(cpm(dge) > 1) >=3
#dge <- dge[isexpr,,keep.lib.sizes=FALSE]

#dge <- calcNormFactors(dge)
```


Calculate log2 CPM
```{r}
cpmNorm <- cpm(dge, normalized.lib.sizes=TRUE, log=TRUE, prior.count=0.25) 
plot.new()
plotDensities(cpmNorm, group = norm.lib.size[colnames(cpmNorm),]$caste) 
abline(v=0)
```

Filter:

if n caste <3 then cpm>1 in n=(n caste)

if n caste >3 then cpm>1 in n=(n caste)/2

```{r}
cpm.dge <- dge
cpm.dge$counts <- cpmNorm

#create a normalization factor 
#if #samples < 3 then norm.factor=1
#if #samples > 3 then norm.factor=0.5

norm.samples <- data.frame(caste=levels(as.factor(cpm.dge$samples$caste)), stringsAsFactors = F)
norm.samples$norm <- ""
isexpr <- unlist(lapply(levels(as.factor(cpm.dge$samples$caste)), function(x){
  if(sum(cpm.dge$samples$caste==x)<3){
    1
  }
  else{
    0.5
  }}))
norm.samples$norm <- isexpr

#the normalization factor is used in
# if n caste <3 then cpm>1 in n=(n caste)
# if n caste >3 then cpm>1 in n=(n caste)/2

#which means that we * by the normalization factor 

#cpm>1
dge.filt1 <- dge[
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="1",])] >1) >= (sum(cpm.dge$samples$caste=="1")*norm.samples$norm[norm.samples$caste=="1"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="2",])] >1) >= (sum(cpm.dge$samples$caste=="2")*norm.samples$norm[norm.samples$caste=="2"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="3",])] >1) >= (sum(cpm.dge$samples$caste=="3")*norm.samples$norm[norm.samples$caste=="3"])
    ,,keep.lib.sizes=FALSE]

cat(dim(dge.filt1)[1], "exons kept in",dim(dge.filt1)[2],"samples if the filter is set to cmp>1\n")

#cmp>0
dge.filt0 <- dge[
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="1",])] >0) >= (sum(cpm.dge$samples$caste=="1")*norm.samples$norm[norm.samples$caste=="1"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="2",])] >0) >= (sum(cpm.dge$samples$caste=="2")*norm.samples$norm[norm.samples$caste=="2"])|
    rowSums(cpm.dge$counts[,rownames(cpm.dge$samples[cpm.dge$samples$caste=="3",])] >0) >= (sum(cpm.dge$samples$caste=="3")*norm.samples$norm[norm.samples$caste=="3"])
    ,,keep.lib.sizes=FALSE]

cat(dim(dge.filt0)[1], "exons kept in",dim(dge.filt0)[2],"samples if the filter is set to cmp>0")

```

Recalculate lib size:

```{r}
dge.filt1 <- calcNormFactors(dge.filt1)
dge.filt0 <- calcNormFactors(dge.filt0)
```

##Voom transformation
Apply the voom transformation and fit a linear model using three normalization methods: 

*No normalization 

*Cyclicloess

*Quantile

Two contrast matrix are used: 

*Caste

*Sex
```{r}

##No normalization
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"No normalization",cex=2,font=2 )
v <- voom(dge.filt0, design, plot=TRUE)

fit <- lmFit(v, design)

v$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v$targets$caste)
v$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v$targets$caste)
v$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v$targets$caste)

contrast.matrix.caste <- makeContrasts(queen-worker, queen-male, worker-male, levels=design)
contrast.matrix.sex <- makeContrasts(male-(queen+worker)/2, levels=design)

fit.caste <- contrasts.fit(fit, contrast.matrix.caste)
fit.caste <- eBayes(fit.caste)

fit.sex <- contrasts.fit(fit, contrast.matrix.sex)
fit.sex <- eBayes(fit.sex)

##Cyclicloess normalization
plot.new()
text(0.5,0.5,"Cyclicloess normalization",cex=2,font=2 )
v.loess <- voom(dge.filt0, design, normalize.method="cyclicloess", plot=TRUE)

fit.loess <- lmFit(v.loess, design)

v.loess$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v.loess$targets$caste)
v.loess$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v.loess$targets$caste)
v.loess$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v.loess$targets$caste)

fit.caste.loess <- contrasts.fit(fit.loess, contrast.matrix.caste)
fit.caste.loess <- eBayes(fit.caste.loess)

fit.sex.loess <- contrasts.fit(fit.loess, contrast.matrix.sex)
fit.sex.loess <- eBayes(fit.sex.loess)

##Quntile normalization
plot.new()
text(0.5,0.5,"Quantile normalization",cex=2,font=2 )
v.quant <- voom(dge.filt0, design, normalize.method="quantile", plot=TRUE)

fit.quant <- lmFit(v.quant, design)

v.quant$targets$caste <- gsub(pattern = "1", replacement = "Queen", x = v.quant$targets$caste)
v.quant$targets$caste <- gsub(pattern = "2", replacement = "Worker", x = v.quant$targets$caste)
v.quant$targets$caste <- gsub(pattern = "3", replacement = "Male", x = v.quant$targets$caste)

fit.caste.quant <- contrasts.fit(fit.quant, contrast.matrix.caste)
fit.caste.quant <- eBayes(fit.caste.quant)

fit.sex.quant <- contrasts.fit(fit.quant, contrast.matrix.sex)
fit.sex.quant <- eBayes(fit.sex.quant)

```

Densities when using lmFit(v, design)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

##Plot densities grouped by caste
plot.new()
text(0.5,0.5,"No normalization",cex=2,font=2 )
plotDensities(v, group = v$targets$caste)

plot.new()
text(0.5,0.5,"Cyclicloess normalization",cex=2,font=2 )
plotDensities(v.loess, group = v$targets$caste)

plot.new()
text(0.5,0.5,"Quantile normalization",cex=2,font=2 )
plotDensities(v.quant, group = v$targets$caste)

```

Densities when using contrasts on caste

contrasts.fit(fit, contrast.matrix.caste)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste",cex=2,font=2 )
plotDensities(fit.caste)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste Cyclicloess norm.",cex=2,font=2 )
plotDensities(fit.caste.loess)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by caste quantile norm.",cex=2,font=2 )
plotDensities(fit.caste.quant)
```

Densities when using contrasts on sex

contrasts.fit(fit, contrast.sex)
```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex",cex=2,font=2 )

plotDensities(fit.sex)
par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex Cyclicloess norm.",cex=2,font=2 )
plotDensities(fit.sex.loess)

par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))

plot.new()
text(0.5,0.5,"Contrast by sex quantile norm.",cex=2,font=2 )
plotDensities(fit.sex.quant)
```

```{r}
as.data.frame(contrast.matrix.caste)
```


```{r}
# par(mar=c(1,1,1,1))
# layout( matrix(c(1,2)), heights = c(1,3))
# 
# plot.new()
# text(0.5,0.5,"MDS Cyclicloess norm.",cex=2,font=2 )
# plotMDS(v.loess,pch = 16, col=as.numeric(as.factor(v.loess$targets$caste)))
# 
# plot.new()
# text(0.5,0.5,"MDS no norm.",cex=2,font=2 )
# plotMDS(v,pch = 16, col=as.numeric(as.factor(v$targets$caste)))
# 
# 
# plot.new()
# text(0.5,0.5,"MDS quantile norm.",cex=2,font=2 )
# plotMDS(v.quant,pch = 16, col=as.numeric(as.factor(v.quant$targets$caste)))

plot.new()
#text(0.5,0.5,"MDS no norm.",cex=2,font=2 )
plotMDS(v.loess,pch = 16, col=as.numeric(as.factor(v$targets$caste)))
legend("topleft", legend=c(levels(as.factor(v$targets$caste))), text.col =c(1,2,3), pch=16)

```

##Differential expression
###Differentially spliced exons by caste

Now we can test for differential splicing associated with any coeficient in the linear model. 

First run the diffSplice function.
```{r results="hold"}
#ex <- diffSplice(fit, geneid="GeneID")
print("No normalization")
ex.caste <- diffSplice(fit.caste, geneid="GeneID")
print("Cyclonloess normalization")
ex.caste.loess <- diffSplice(fit.caste.loess, geneid="GeneID")
print("Quantile normalization")
ex.caste.quant <- diffSplice(fit.caste.quant, geneid="GeneID")
```

Will find genes that show evidence of differential splicing associated with the first (Queen-Worker) coefficient in the linear model. 

The output is similar that from the limma topTable function. 

```{r rows.print=10}
#More detail can be obtained by topSplice

#First define the number of differentially expressed exons to show
numDiff.caste <- 200000
#then use the function topSplice with the coeficient (comparision) that we're interested in
diff.caste <- topSplice(ex.caste.loess, coef=1, test="simes", number=numDiff.caste,  FDR = 0.5)
diff.caste$mstrgid <- sub(".*(MSTRG\\.\\d*).*", "\\1", diff.caste$GeneID)
dplyr::select(diff.caste, genes, GeneID, mstrgid)

#Find the names of the genes is Drosophila 
diffMatches.caste <- sapply(diff.caste$mstrgid, function(x){matchList[[x]]})
as.data.frame(unlist(diffMatches.caste))

#Find out how many differentially splices exons in the non model have
# 0 matches in the model species genome
# 1 match in the model species genome
# >2 matches in the model species genome

noMatch.caste <- sum(sapply(sapply(diff.caste$mstrgid, function(x){matchList[[x]]}), function(x){ length(x)  ==0} ) )
oneMatch.caste <- sum(sapply(sapply(diff.caste$mstrgid, function(x){matchList[[x]]}), function(x){ length(x)  ==1} ) )
multipleMatch.caste <- sum(sapply(sapply(diff.caste$mstrgid, function(x){matchList[[x]]}), function(x){ length(x)  >=2} ) )
```

```{r results="asis"}
#Plot the number of matches of exons of the non-model organism to the model 

cat("From", noMatch.caste+oneMatch.caste+multipleMatch.caste ,"differentially spliced genes:\n", noMatch.caste, "have no match in the model species\n", oneMatch.caste, "have 1 match in the model species\n", multipleMatch.caste, "have >2 match in the model species\n")


matches.summary.caste <- data.frame(Diff_spliced_exons=c(noMatch.caste, oneMatch.caste, multipleMatch.caste), match=factor(c("0", "1", ">2"), levels =c("0", "1", ">2")))

ggplot(matches.summary.caste, aes(match, Diff_spliced_exons))+
  geom_bar(stat="identity")

```

```{r}
#topSplice(ex, coef=1, test="t", number = 200)
percentMapped=100-1715/2734*100
# percentMapped=100-((noMatch.caste)/(noMatch.caste+oneMatch.caste+multipleMatch.caste)*100)
cat("Percentage of DS genes mapped to Drosophila melanogaster:", percentMapped)
```

###Differential splicing of _doublesex_ by caste

Which will show individual exons that are enriched or depleted relative to other exons in the same gene. 

To display the pattern of exons in the top genes.

Plot doublesex (identified by blast beforehand)  "gene13255:MSTRG.15080"

```{r}
par(mar=c(1,1,1,1))
layout( matrix(c(1,2,3,4,5,6),ncol=1), heights = rep(c(1,3), 3))
#layout.show(n = 6)
plot.new()
text(0.5,0.5,"Queen-Worker",cex=2,font=2 )
plotSplice(ex.caste.loess, coef = 1, geneid = "gene13255:MSTRG.15080")

plot.new()
text(0.5,0.5,"Queen-Male",cex=2,font=2)
plotSplice(ex.caste.loess, coef = 2, geneid = "gene13255:MSTRG.15080")

plot.new()
text(0.5,0.5,"Worker-Male",cex=2,font=2)
plotSplice(ex.caste.loess, coef = 3, geneid = "gene13255:MSTRG.15080")

#dev.off()

```

P values of the exons per contrast

```{r}

#diff$genes=="gene13255:MSTRG.15080"
pvalues_dsx <- ex.caste.loess$p.value[ex.caste.loess$gene.firstexon["gene13255:MSTRG.15080"]:ex.caste.loess$gene.lastexon["gene13255:MSTRG.15080"],]
#rownames(pvalues_dsx) <- c("exon1","exon3","exon5","exon6","exon7")
pvalues_dsx
ex.caste.loess[ex.caste.loess$gene.firstexon["gene13255:MSTRG.15080"]:ex.caste.loess$gene.lastexon["gene13255:MSTRG.15080"],]
dge[dge$genes$gene=="gene13255:MSTRG.15080",]
```


###Differentially spliced exons by sex

Now we can test for differential splicing associated with any coeficient in the linear model. 

First run the diffSplice function.
```{r results="hold"}
#ex <- diffSplice(fit, geneid="GeneID")
ex.sex <- diffSplice(fit.sex.loess, geneid="GeneID")
```

Will find genes that show evidence of differential splicing associated with the second coefficient in the linear model. 

The output is similar that from the limma topTable function. 

```{r results="hold"}
#More detail can be obtained by
numDiff.sex <- 200000
diff.sex <- topSplice(ex.sex, coef=1, test="simes", number=numDiff.caste,  FDR = 0.5)
diff.sex$mstrgid <- sub(".*(MSTRG\\.\\d*).*", "\\1", diff.sex$GeneID)
dplyr::select(diff.sex, genes, GeneID, mstrgid)

#Find the names of the genes is Drosophila 
diffMatches.sex <- sapply(diff.sex$mstrgid, function(x){matchList[[x]]})
as.data.frame(unlist(diffMatches.sex))

#Find out how many differentially splices exons in the non model have
# 0 matches in the model species genome
# 1 match in the model species genome
# >2 matches in the model species genome

noMatch.sex <- sum(sapply(sapply(diff.sex$mstrgid, function(x){matchList[[x]]}), function(x){ length(x)  ==0} ) )
oneMatch.sex <- sum(sapply(sapply(diff.sex$mstrgid, function(x){matchList[[x]]}), function(x){ length(x)  ==1} ) )
multipleMatch.sex <- sum(sapply(sapply(diff.sex$mstrgid, function(x){matchList[[x]]}), function(x){ length(x)  >=2} ) )
```

```{r results="asis"}
#Plot the number of matches of exons of the non-model organism to the model 

cat("From", noMatch.sex+oneMatch.sex+multipleMatch.sex ,"differentially spliced genes:\n", noMatch.sex, "have no match in the model species\n", oneMatch.sex, "have 1 match in the model species\n", multipleMatch.sex, "have >2 match in the model species\n")


matches.summary.sex <- data.frame(Diff_spliced_exons=c(noMatch.sex, oneMatch.sex, multipleMatch.sex), match=factor(c("0", "1", ">2"), levels =c("0", "1", ">2")))

ggplot(matches.summary.sex, aes(match, Diff_spliced_exons))+
  geom_bar(stat="identity")

```

```{r}
#topSplice(ex, coef=1, test="t", number = 200)
```

###Differential splicing of _doublesex_ by sex

Which will show individual exons that are enriched or depleted relative to other exons in the same gene. 

To display the pattern of exons in the top genes.

Plot doublesex (identified by blast beforehand)  "gene13255:MSTRG.15080"

```{r}


par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))
#layout.show(n = 2)
plot.new()
text(0.5,0.5,"Male-Female",cex=2,font=2 )
plotSplice(ex.sex, coef = 1, geneid = "gene13255:MSTRG.15080")

#dev.off()

```

P values of the exons per contrast

```{r}

#diff$genes=="gene13255:MSTRG.15080"
pvalues_dsx.sex <- ex.sex$p.value[ex.sex$gene.firstexon["gene13255:MSTRG.15080"]:ex.sex$gene.lastexon["gene13255:MSTRG.15080"],]
#names(pvalues_dsx.sex) <- c("exon1","exon2","exon3","exon4","exon5","exon6","exon7")
pvalues_dsx.sex
```

###Intersect Sex and Caste

```{r}

sex_genes <- data.frame(gene=unlist(diffMatches.sex), contrast=rep("sex", length(unlist(diffMatches.sex))))
caste_genes <- data.frame(gene=unlist(diffMatches.caste), contrast=rep("caste", length(unlist(diffMatches.caste))))
matchedgenes <- rbind(caste_genes, sex_genes)

grid.newpage()
venn.plot <- draw.pairwise.venn(area1 = length(unlist(diffMatches.sex)), 
                   area2 = length(unlist(diffMatches.caste)), 
                   cross.area = length(intersect((unlist(diffMatches.sex)), (unlist(diffMatches.caste)))), 
                   category = c("DS sex", "DS caste"))

diffOneMatch.caste <- diffMatches.caste[sapply(diffMatches.caste, function(x){length(x)==1})]
diffOneMatch.sex <- diffMatches.sex[sapply(diffMatches.sex, function(x){length(x)==1})]

grid.newpage()
venn.plot <- draw.pairwise.venn(area1 = length(unlist(diffOneMatch.sex)), 
                   area2 = length(unlist(diffOneMatch.caste)), 
                   cross.area = length(intersect((unlist(diffOneMatch.sex)), (unlist(diffOneMatch.caste)))), 
                   category = c("DS sex", "DS caste"))

#Save the data with names specific to the species (to do a cross-species comparison)
Vem_diffMatches.sex <- diffMatches.sex
Vem_diffMatches.caste <- diffMatches.caste
Vem_intersect_diffMatches <- intersect((unlist(diffMatches.sex)), (unlist(diffMatches.caste)))

write.table(x = as.data.frame(unlist(Vem_diffMatches.sex)), file = "../../R/Vem_diffMatches_sex.txt", col.names = TRUE, row.names = TRUE)
write.table(x = as.data.frame(unlist(Vem_diffMatches.caste)), file = "../../R/Vem_diffMatches_caste.txt", col.names = TRUE, row.names = TRUE)
```

```{r}
#load Apis mellifera gff 
vem_gtf <- import.gff3("/Users/afarre/Unimelb/Doublesex/Vemeryi/genome/GCF_000949405.1_V.emery_V1.0_genomic.gff")


n <- 200000
nshort <- 20
a <- topSplice(ex.caste, coef=1, test="simes", number=n,  FDR = 0.5)
dcaste <- sapply(unlist(strsplit(a$GeneID, ":"))[grep("gene*", unlist(strsplit(a$GeneID, ":")))], function(e){vem_gtf[mcols(vem_gtf)$ID==e]$Name})

nay <- na.omit(as.data.table(vem_gtf), cols="gene")
prod.dcaste <- lapply(dcaste, function(e){nay[nay$gene==e]$product[2]})

prod.dcaste.df <- as.data.frame(unlist(prod.dcaste))
names(prod.dcaste.df)<-"product"

prod.dcaste.df$product <- sub("(.*), transcript variant .*", "\\1", prod.dcaste.df$product)

prod.dcaste.df <- unique(prod.dcaste.df)
prod.dcaste.df$geneNumber <- rownames(prod.dcaste.df)
int_genes <- prod.dcaste.df[grep(".*transformer.*", prod.dcaste.df$product, ignore.case = TRUE),]
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*vitellogenin.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*bric.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*juvenile hormone.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*insulin.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*royal.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*forkhead.*O-like.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*yellow.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*doublesex.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*fruitless.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*abrupt-like.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*sex-lethal.*", prod.dcaste.df$product, ignore.case = TRUE),])
int_genes <- rbind(int_genes, prod.dcaste.df[grep(".*feminizer.*", prod.dcaste.df$product, ignore.case = TRUE),])

write.table(x = int_genes, file = "/Users/afarre/Unimelb/Doublesex/R/Vem_prepupa_AS_int_genes.txt", col.names = TRUE, row.names = TRUE)
int_genes
```

```{r}

targetgene <- "gene8677:"
targetgene <- diff.caste[grep(targetgene, diff.caste$genes),]$genes

par(mar=c(1,1,1,1))
layout( matrix(c(1,2,3,4,5,6),ncol=1), heights = rep(c(1,3), 3))
#layout.show(n = 6)
plot.new()
text(0.5,0.5,"Queen-Worker",cex=2,font=2 )
plotSplice(ex.caste.loess, coef = 1, geneid = targetgene)

plot.new()
text(0.5,0.5,"Queen-Male",cex=2,font=2)
plotSplice(ex.caste.loess, coef = 2, geneid = targetgene)

plot.new()
text(0.5,0.5,"Worker-Male",cex=2,font=2)
plotSplice(ex.caste.loess, coef = 3, geneid = targetgene)

```

```{r}


par(mar=c(1,1,1,1))
layout( matrix(c(1,2)), heights = c(1,3))
#layout.show(n = 6)
plot.new()
text(0.5,0.5,"Male-Female",cex=2,font=2 )
plotSplice(ex.sex, coef = 1, geneid = targetgene)


```
