---
title: "Wau_necklace_DEXSeq_NoFilter"
author: "Anna Orteu"
date: "20/06/2018"
output: 
  html_document:
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 4  # upto 4 depths of headings (specified by #, ##, ### and ####)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: flatly
    code_folding: hide # Add some awesome buttons to show/hide the code
    df_print: paged
---

Script to analyse necklace output using DEXSeq. 
Saves results in tables. 

Input requited: 

* block.counts file from necklace

* X_SraRunTable.txt file with information about the samples 

#Prepare workspace 

Load libraries
```{r message= FALSE, warning=FALSE}
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("tidyr")
#devtools::install_github("sjmgarnier/viridis")
#install.packages("VennDiagram")
#install.packages("rtracklayer")

#source("https://bioconductor.org/biocLite.R")
#biocLite("limma")
#biocLite("edgeR")
#biocLite("DEXSeq")
#biocLite("GenomicRanges")
#biocLite("Rsubread")



library(limma)
library(edgeR)
library(ggplot2)
library(dplyr)
library(plyr)
library(tidyr)
library(DEXSeq)
library(GenomicRanges)
library(rtracklayer)
library(Rsubread)
library(viridis) #colour blind friendly palette for plots
library(VennDiagram)
library(rtracklayer)
library(dplyr)
library(data.table)
library(gplots)
library(UpSetR)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Differential Splicing: Analysis of necklace output of _Monomorium pharaonis_ 
##Pre-processing
###Read in the data info

DexSeq requires a table with information about the samples as well as the raw counts for each exon in each sample
And grouping information for which exons belong to which genes

Read in the table with all the information of the SRA samples

```{r rows.print=5}
sample_id<-read.table("/Users/afarre/Unimelb/Doublesex/Wauropunctatat/data_info/Wau_DRP002449_SRR_Acc_List.txt", col.names = "sampleID", stringsAsFactors = F )

#exclude a pooled sample
sample_id <- filter(sample_id, sampleID!="DRR029092")

info_samples<-read.table("/Users/afarre/Unimelb/Doublesex/Wauropunctatat/data_info/Wau_DRP002449_SraRunTable.txt", sep = "\t", header = T,  stringsAsFactors = F )
info_samples <- filter(info_samples, Run_s!="DRR029092")
rownames(info_samples) <- info_samples$Run_s

info_samples
```

Create the data_info table for DEXseq

```{r rows.print=10}
info_samples$caste <- "1"
info_samples[grep("queen", info_samples$sample_title_s),]$caste <- "1"
info_samples[grep("worker", info_samples$sample_title_s),]$caste <- "2"
info_samples[grep("male", info_samples$sample_title_s),]$caste <- "3"

info_samples$casteName <- ""
info_samples[grep("queen", info_samples$sample_title_s),]$casteName <- "queen"
info_samples[grep("worker", info_samples$sample_title_s),]$casteName <- "worker"
info_samples[grep("male", info_samples$sample_title_s),]$casteName <- "male"

info_samples$population <- ""
info_samples[grep("French Guiana", info_samples$sample_title_s),]$population <- "1"
info_samples[grep("hawai", info_samples$sample_title_s),]$population <- "2"
row.names(info_samples) <- info_samples$Run_s

info_samples$caste <- as.factor(info_samples$caste)
info_samples$population <- as.factor(info_samples$population)


data_info <- dplyr::select(info_samples, Run_s, casteName, population)
row.names(data_info) <- data_info$Run_s
data_info$casteName <- as.factor(data_info$casteName)
colnames(data_info) <- c("sample", "condition", "population")

```


###Read in the count data
Read in the data (count matrix)
```{r  rows.print=5}
#First get counts table
counts <- read.table("/Users/afarre/Unimelb/Doublesex/Wauropunctatat/necklace_ame/block.counts",sep="\t",header=TRUE,stringsAsFactors=F)
names(counts) <- gsub(".*(DRR.*)\\.bam","\\1", names(counts))
counts <- dplyr::select(counts, -DRR029092)

#select only the samples that we want to use
#in this case it's only the PE samples
counts <- dplyr::select(counts, Geneid, Chr, Start, End, Strand, Length, data_info$sample)
counts$Geneid <- gsub(":","_",counts$Geneid)

```

Create the exon_ids column with a unique identifier for each exon

```{r rows.print=5}
#each exon needs a unique name
#create a column with unique exon ids 
#exonid = geneid + exon number
counts <- do.call(rbind, lapply(unique(counts$Geneid), function(x){ 
  mutate(dplyr::filter(counts,Geneid==x), exon_ids=
           paste(dplyr::filter(counts,Geneid==x)$Geneid, 
                 rownames(dplyr::filter(counts,Geneid==x)), sep = "."))
} ))

row.names(counts) <- counts$exon_ids
```

Convert the counts data frame into a matrix (i.e exclude first 6 columns: gene, chr, etc.)

```{r}
#create the matrix with only the counts
count_matrix <- data.matrix(dplyr::select(counts, -Geneid, -Chr, -Start, -End, -Strand, -Length, -exon_ids))
count_matrix <- round(count_matrix)
# gene_ids <- counts$Geneid
# exon_ids <- counts$exon_ids
```

##DEXSeq Queens vs Workers

Small testing dataset
```{r}
data_info %>%
  filter(condition!="male") %>%
  dplyr::select(sample)

# Small dataset to test DEXSeq models
# nshort <- 1000
# dgeQW <- head(count_matrix[,data_info[data_info$condition!="male",]$sample], nshort)
# 
# gene_ids <- head(counts$Geneid, nshort)
# exon_ids <- head(counts$exon_ids, nshort)
```

Full dataset
```{r}
#Full dataset
dgeQW <- count_matrix[,data_info[data_info$condition!="male",]$sample]

gene_ids <- counts$Geneid
exon_ids <- counts$exon_ids


#The model
formulaFullModel = ~ sample + exon + population + condition:exon
formulaReducedModel = ~ sample + exon + population

#Create the dexseq object
dxd <- DEXSeqDataSet(dgeQW,design=formulaFullModel,featureID=as.factor(exon_ids),
                     groupID=as.factor(gene_ids),sampleData=data_info[data_info$condition!="male",])

## Could use the DEXSeqDataSetFromFeatureCounts function, but never got it to work...
# flattenedFile <- import.gff("/Users/afarre/Unimelb/Doublesex/Mpharaonis/necklace_ame/SuperDuper.gff")
# source("/Users/afarre/Unimelb/Doublesex/R/load_SubreadOutput.R")
# dxd2 <- DEXSeqDataSetFromFeatureCounts(countfile = "/Users/afarre/Unimelb/Doublesex/Mpharaonis/necklace_ame/block_counts.out" ,design=~sample + exon + condition:exon, sampleData=data_info, flattenedfile="/Users/afarre/Unimelb/Doublesex/Mpharaonis/necklace_ame/SuperDuper.gff")


#Can do some pre-filtering that will speed up the process and reduce the number of tests 
#hard filtering is automatically aplied by DEXSeq
keep <- rowSums(counts(dxd)) >= 10
dxd <- dxd[keep,]

#Estimate size factors and dispersions (DEXseq does this based on a negative bionmial distribution
dxd <- estimateSizeFactors(dxd)
dxd = estimateDispersions( dxd, formula = formulaFullModel)

#plot dispersions
#A good result has a curve, decreasing fitted (red) line
plotDispEsts(dxd)

#Test for DEU
dxd = testForDEU( dxd,reducedModel = formulaReducedModel,fullModel = formulaFullModel )

#Extract the results
res <- DEXSeqResults(dxd)

#Get p-value per gene based on the p-values for all exons in the gene/cluster
pgq <- perGeneQValue(res, p = "pvalue" ) 
pgq

## Save results to a text file and R object
tmp <- cbind(gene = names(pgq), "adjP" = pgq)
```

Save the Rdata and tables 

```{r}
save(dxd, res, pgq, file = "/Users/afarre/Unimelb/Doublesex/R/DEXSeq/Wau_DEXseqQW.Rdata")
write.table(tmp, file = "/Users/afarre/Unimelb/Doublesex/R/DEXSeq/Wau_DEXseq_byGeneQW.txt", col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(DEXSeqResults(dxd), file="/Users/afarre/Unimelb/Doublesex/R/DEXSeq/Wau_DEXseq_byExonQW.txt", col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t" )
```
